% Kaappi: A Scheme Programming Language
% Compiled with XeLaTeX

\documentclass[11pt,twoside,openright]{memoir}

% Load preamble files
\input{preamble/colors}
\input{preamble/packages}
\input{preamble/fonts}
\input{preamble/fonts-memoir}
\input{preamble/decorations}
\input{preamble/layout}
\input{preamble/styles}
\input{preamble/listings}

% Kaappi version shown in prose and transcripts. Keep in sync with releases;
% inside listings, expand it with the (*@\kaappiversion@*) escape.
\newcommand{\kaappiversion}{0.11.0}

\begin{document}

% Front matter
\frontmatter
\pagestyle{empty}

% Title page
\thispagestyle{empty}
\begin{center}
    \vspace*{3cm}
    {\fontsize{32}{38}\selectfont\sffamily\bfseries\color{darkbrown}
     Kaappi}\\[1cm]
    {\Large\sffamily A Scheme Programming Language}\\[3cm]
    {\large\sffamily Baiju Muthukadan}\\[1cm]
    \vfill
    {\normalsize\sffamily R7RS-small $\bullet$ Zig $\bullet$ Batteries Included}\\[1.5cm]
\end{center}
\newpage

% Copyright page
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
    \textit{Kaappi: A Scheme Programming Language}\\[1em]
    Baiju Muthukadan\\[2em]
    First Edition, 2026\\[2em]
    Copyright \textcopyright{} 2026 Baiju Muthukadan\\
    All rights reserved.\\[2em]
    \vspace{2em}
    \textbf{Draft Edition}\\[0.5em]
    This book is a work in progress.\\
    Feedback is welcome at \href{mailto:baiju.m.mail@gmail.com}{baiju.m.mail@gmail.com}.
\end{center}
\vspace*{\fill}
\cleardoublepage

% Table of Contents
\pagestyle{plain}
\tableofcontents*
\cleardoublepage

% Preface
\chapter{Preface}

\section*{Why Scheme?}

If you already know Python, JavaScript, or Ruby, you can write programs that work. So why learn another language---especially one with all those parentheses?

Scheme\index{Scheme} will change the way you think about programming. Its syntax is so minimal that you can hold the entire language in your head. There are no special forms for loops, no operator precedence rules, no class hierarchies. Instead, you get a small set of powerful ideas---first-class functions, lexical closures, recursion, and macros---that combine to express anything you need.

This matters because Scheme's ideas are everywhere. Python borrowed \texttt{lambda}, list comprehensions, and garbage collection from Lisp. JavaScript's closures and first-class functions come directly from Scheme. Ruby's blocks are a limited form of Scheme's closures. Learning Scheme means understanding the foundations that these languages are built on.

\section*{Why Kaappi?}

Kaappi\index{Kaappi} is a modern, complete implementation of the R7RS-small\index{R7RS} standard. Unlike many Scheme implementations that focus on either academic purity or raw performance, Kaappi aims to be a practical, batteries-included language for building real programs.

It ships as a statically linked binary with no shared library dependencies---just the binary and a library archive of portable Scheme files. The interactive REPL\index{REPL} provides syntax highlighting, tab completion, and history---a comfortable environment for exploration. Beyond the standard, Kaappi includes 72 SRFIs\index{SRFI} (8 compiled into the binary, 64 as portable Scheme libraries), a C foreign function interface, green threads with channels, a package manager, and an ecosystem of libraries for web servers, databases, JSON, and more.

If you have ever tried a Scheme and felt it was ``just a toy,'' Kaappi is different. You can build real applications with it.

\section*{Who This Book Is For}

This book is for programmers who know at least one mainstream language---Python, JavaScript, Ruby, Java, Go, or similar---and want to learn Scheme. No prior experience with Lisp or functional programming is required.

This is not an academic textbook. You will not find formal semantics or denotational models here. Instead, you will find clear explanations, running code, and practical exercises. Theory is introduced only when it helps you write better programs.

\section*{What You Will Need}

\begin{itemize}
    \item \textbf{Kaappi installed on your system.} Chapter~2 covers installation in detail. The short version: \texttt{curl -fsSL https://kaappi-lang.org/\allowbreak install.sh | bash}
    \item \textbf{A terminal.} You will spend most of your time in the Kaappi REPL, typing expressions and seeing results immediately.
    \item \textbf{A text editor.} Any editor works. If you use VS Code, Neovim, or Emacs, Kaappi provides an LSP server for go-to-definition, find-references, and document symbols.
\end{itemize}

\section*{How to Use This Book}

The book is organized into four parts:

\begin{description}[leftmargin=1em, labelindent=0em]
    \item[Part I: Getting Started] covers installation, the REPL, and basic expressions. Start here.
    \item[Part II: The Language] teaches the core of Scheme: data types, lists, control flow, functions, bindings, higher-order functions, and I/O. This is the heart of the book.
    \item[Part III: Going Further] introduces macros, the library system, records, error handling, and continuations. Read this once you are comfortable writing functions and working with lists.
    \item[Part IV: Beyond R7RS] explores Kaappi-specific features: the SRFI ecosystem, the C foreign function interface, and concurrency with fibers and threads.
\end{description}

Each chapter builds on the previous ones. If you are new to Scheme, work through them in order. If you already know some Scheme and want to learn what Kaappi offers beyond the standard, skip to Part~IV.

The appendices are reference material to dip into as needed: an R7RS quick reference, a command-line reference, the full SRFI catalog, common error messages and their fixes, the ecosystem package list, a glossary, and pointers for further reading. If you hit an unfamiliar term while reading, the Glossary (Appendix~\ref{app:glossary}) is the fastest place to look it up.

The most effective way to use this book is with the Kaappi REPL open beside it. Type every example. Modify them. Break them. The REPL gives instant feedback, and that tight loop between reading and trying is how Scheme clicks.

Exercises appear at the end of most chapters. They are not optional filler---they are where you solidify your understanding. Attempt them before moving on.

This book does not print exercise solutions, and that is deliberate. Scheme makes it easy to check your own work: type your answer into the REPL, compare its output against the expected result shown in the exercise, and refine until they match. Many exercises are open-ended---there is no single correct answer, and the point is to build fluency rather than match a key. If you get truly stuck, reread the relevant section with the REPL open and experiment until the behavior makes sense.

\section*{Conventions}

Throughout this book:

\begin{description}[leftmargin=1em, labelindent=0em]
    \item[Code listings] show Scheme source code with syntax highlighting. Keywords are bold.
    \item[REPL sessions] show interactive input (lines starting with \texttt{>}) and output. Type the input lines into your Kaappi REPL to follow along.
    \item[Notes] provide supplementary information, tips, or comparisons to Python.
    \item[Warnings] highlight common mistakes and pitfalls.
    \item[Summaries] recap a chapter's key points in a short bulleted list at the end, just before the exercises.
    \item[Exercises] are numbered per chapter and provide practice problems of increasing difficulty.
\end{description}

\cleardoublepage

% Main matter
\mainmatter
\pagestyle{kaappibook}

% Part I: Getting Started
\part{Getting Started}

This part takes you from nothing to a working setup. You will see what Scheme and Kaappi are, install the interpreter, get comfortable in the REPL, and write your first expressions and functions. By the end you will be ready to explore the language in depth.

\input{chapters/ch01-introduction}
\input{chapters/ch02-installation}
\input{chapters/ch03-first-steps}

% Part II: The Language
\part{The Language}

This is the heart of the book. These chapters cover the core of Scheme---data types, lists and pairs, control flow, functions, local bindings, higher-order functions, and input and output---introducing one concept at a time. Work through them in order, since each chapter builds directly on the ones before it.

\input{chapters/ch04-data-types}
\input{chapters/ch05-lists-and-pairs}
\input{chapters/ch06-control-flow}
\input{chapters/ch07-functions}
\input{chapters/ch08-local-bindings}
\input{chapters/ch09-higher-order-functions}
\input{chapters/ch10-input-and-output}

% Part III: Going Further
\part{Going Further}

With the core language in hand, this part covers the tools you reach for as programs grow: macros that extend Scheme's own syntax, the library system for organizing code into modules, user-defined record types, structured error handling, and continuations. Read it once you are comfortable writing functions and working with lists.

\input{chapters/ch11-macros}
\input{chapters/ch12-library-system}
\input{chapters/ch13-records}
\input{chapters/ch14-error-handling}
\input{chapters/ch15-continuations}

% Part IV: Beyond R7RS
\part{Beyond R7RS}

This part explores what Kaappi adds on top of the standard: its catalog of SRFI libraries, the C foreign function interface for calling native code, and concurrency with fibers and threads. If you already know Scheme, this is a good place to start.

\input{chapters/ch16-standard-library}
\input{chapters/ch17-ffi}
\input{chapters/ch18-concurrency}

% Appendices
\appendix
\part{Appendices}

The appendices are reference material to reach for as you read or keep handy while you program: a quick R7RS reference, the command-line interface, the full SRFI catalog, common error messages, the ecosystem packages, a glossary, and pointers for further study.

\input{chapters/appendix-a-r7rs-reference}
\input{chapters/appendix-b-cli-reference}
\input{chapters/appendix-c-srfi-catalog}
\input{chapters/appendix-d-error-messages}
\input{chapters/appendix-e-ecosystem}
\input{chapters/appendix-f-glossary}
\input{chapters/appendix-g-further-reading}

% Back matter
\backmatter

\printindex

\end{document}
