\chapter{Further Reading}
\index{further reading}
\label{app:reading}

This appendix collects books, standards documents, and online resources for continuing your study of Scheme and Kaappi.

\section{Books}

\begin{description}[leftmargin=1em, labelindent=0em]

\item[Structure and Interpretation of Computer Programs] \textit{Harold Abelson, Gerald Jay Sussman, and Julie Sussman} (MIT Press, 1996). The classic introduction to computer science using Scheme. Covers abstraction, recursion, interpreters, and compilation. Available free online from MIT Press.

\item[The Scheme Programming Language, 4th Edition] \textit{R.\ Kent Dybvig} (MIT Press, 2009). A thorough reference and tutorial for the Scheme language. Covers R6RS but most content applies to R7RS. Available free online at \url{https://scheme.com/tspl4/}.

\item[The Little Schemer] \textit{Daniel P.\ Friedman and Matthias Felleisen} (MIT Press, 1996). A short, playful introduction to recursive thinking using Scheme. Structured as a series of questions and answers that build up to writing a Scheme interpreter in Scheme. Highly recommended if you want to deepen your intuition for recursion.

\item[The Seasoned Schemer] \textit{Daniel P.\ Friedman and Matthias Felleisen} (MIT Press, 1996). The sequel to \textit{The Little Schemer}. Covers continuations, mutation, and advanced control flow.

\item[How to Design Programs, 2nd Edition] \textit{Matthias Felleisen, Robert Bruce Findler, Matthew Flatt, and Shriram Krishnamurthi} (MIT Press, 2018). A systematic introduction to program design using a Scheme-family teaching language. Gentler than SICP. Available free online at \url{https://htdp.org}.

\item[Essentials of Programming Languages, 3rd Edition] \textit{Daniel P.\ Friedman and Mitchell Wand} (MIT Press, 2008). Uses Scheme to build interpreters for progressively more complex languages. Excellent for understanding how programming languages work.

\end{description}

\section{Standards and Specifications}

\begin{description}[leftmargin=1em, labelindent=0em]

\item[R7RS-small Report] The official language standard that Kaappi implements. Defines all syntax forms, standard libraries, and semantics. Available at \url{https://r7rs.org}. Essential reading once you are comfortable with the language. A follow-on effort, R7RS-large, is standardizing a much larger library set on top of it.

\item[SRFI Process] Scheme Requests for Implementation---the community process for proposing and standardizing library extensions. Browse all SRFIs at \url{https://srfi.schemers.org}. Each SRFI includes a rationale, specification, and reference implementation. Appendix~\ref{app:srfi} catalogs the 85 SRFIs Kaappi ships.

\end{description}

\section{Kaappi Resources}

\begin{description}[leftmargin=1em, labelindent=0em]

\item[Documentation Site] \url{https://kaappi-lang.org} --- the complete Kaappi reference, including the guided tutorial, language quick reference, cookbook, and procedure documentation organized by category (600+ procedures).

\item[Tutorial] \url{https://kaappi-lang.org/guide/tutorial/} --- a 12-section beginner tutorial that complements this book.

\item[Tour] \url{https://kaappi-lang.org/tour/} --- 12 interactive lessons that run live code in your browser. A quick way to refresh a concept without leaving the page.

\item[Cookbook] \url{https://kaappi-lang.org/cookbook/} --- practical recipes for common tasks: building REST APIs, parsing JSON, working with databases, writing CLI tools, and testing.

\item[Playground] \url{https://kaappi-lang.org/playground/} --- a browser-based REPL running Kaappi via WebAssembly. Useful for quick experiments without a local install.

\item[Community] \url{https://kaappi-lang.org/community/} --- where to ask questions, report bugs, and follow development.

\item[Source Code] \url{https://github.com/kaappi/kaappi} --- the Kaappi implementation in Zig. Reading the source is the best way to understand how the language works at a deep level.

\end{description}

\section{Online Resources}

\begin{description}[leftmargin=1em, labelindent=0em]

\item[Scheme Community Wiki] \url{https://community.schemewiki.org} --- tutorials, implementation comparisons, and community-maintained documentation.

\item[Awesome Scheme] \url{https://github.com/schemedoc/awesome-scheme} --- a curated list of Scheme implementations, libraries, and resources.

\item[Scheme and Lisp Communities] \url{https://reddit.com/r/scheme} and \url{https://reddit.com/r/lisp} --- active forums for discussing Scheme and Lisp topics.

\end{description}
