\chapter{Installation and Setup}
\index{installation}
\label{ch:installation}

This chapter gets Kaappi running on your system, introduces the interactive REPL, and shows you how to run script files. By the end, you will have a working environment for the rest of the book.

\section{Installing Kaappi}
\index{install script}

\subsection{Install Script (Recommended)}

The fastest way to install Kaappi is the one-line install script. It detects your platform, downloads the correct binary, verifies SHA256 checksums, and places everything in the right location:

\begin{Verbatim}[fontsize=\small]
curl -fsSL https://kaappi-lang.org/install.sh | bash
\end{Verbatim}

This installs two binaries to \texttt{\textasciitilde/.local/bin/}:

\begin{itemize}
    \item \texttt{kaappi} --- the Scheme interpreter, compiler, and REPL.
    \item \texttt{thottam}\index{thottam} --- the package manager for installing third-party libraries.
\end{itemize}

Standard libraries (SRFIs and Scheme base libraries) are extracted to \texttt{\textasciitilde/.kaappi/lib/}.

To install to a different location, set the \texttt{INSTALL\_DIR} environment variable:

\begin{Verbatim}[fontsize=\small]
curl -fsSL https://kaappi-lang.org/install.sh \
  | INSTALL_DIR=/usr/local/bin bash
\end{Verbatim}

\begin{warning}[PATH configuration]
Make sure \texttt{\textasciitilde/.local/bin} is in your shell's \texttt{PATH}. If running \texttt{kaappi} after installation gives ``command not found,'' add this line to your \texttt{\textasciitilde/.bashrc}, \texttt{\textasciitilde/.zshrc}, or equivalent:

\begin{Verbatim}[fontsize=\small]
export PATH="$HOME/.local/bin:$PATH"
\end{Verbatim}

Then restart your terminal or run \texttt{source \textasciitilde/.zshrc}.
\end{warning}

\subsection{Manual Download}

If you prefer not to pipe a script to your shell, download the binary directly from the GitHub releases page. Kaappi provides pre-built binaries for these platforms:

\begin{center}
\begin{tabular}{lll}
\textbf{OS} & \textbf{Architecture} & \textbf{Binary name} \\
\hline
macOS & ARM64 (Apple Silicon) & \texttt{kaappi-aarch64-macos} \\
Linux & x86\_64 & \texttt{kaappi-x86\_64-linux} \\
Linux & ARM64 & \texttt{kaappi-aarch64-linux} \\
Linux & RISC-V 64 & \texttt{kaappi-riscv64-linux} \\
\end{tabular}
\end{center}

\begin{note}[Windows]
There are no native Windows binaries---Kaappi depends on POSIX APIs. On Windows, install the Linux x86\_64 build inside WSL2 (Windows Subsystem for Linux); everything in this book works there.
\end{note}

Each release also ships a matching \texttt{thottam} package-manager binary per platform (e.g., \texttt{thottam-aarch64-macos})---download it alongside \texttt{kaappi}. After downloading, make the binaries executable and move them to a directory in your PATH:

\begin{Verbatim}[fontsize=\small]
chmod +x kaappi-aarch64-macos thottam-aarch64-macos
mv kaappi-aarch64-macos ~/.local/bin/kaappi
mv thottam-aarch64-macos ~/.local/bin/thottam
\end{Verbatim}

You also need the standard library archive. Download \texttt{kaappi-lib.tar.gz} and extract it:

\begin{Verbatim}[fontsize=\small]
mkdir -p ~/.kaappi/lib
tar xzf kaappi-lib.tar.gz -C ~/.kaappi/lib
\end{Verbatim}

\subsection{Building from Source}
\index{Zig}

Building from source requires Zig 0.16 or later. No other build tools (make, cmake, cargo) are needed.

On macOS:

\begin{Verbatim}[fontsize=\small]
brew install zig
git clone https://github.com/kaappi/kaappi
cd kaappi
zig build
\end{Verbatim}

On Linux, download Zig from \texttt{ziglang.org/download}, add it to your PATH, then:

\begin{Verbatim}[fontsize=\small]
git clone https://github.com/kaappi/kaappi
cd kaappi
zig build
\end{Verbatim}

The binary appears at \texttt{zig-out/bin/kaappi}. Copy it to a location in your PATH:

\begin{Verbatim}[fontsize=\small]
cp zig-out/bin/kaappi zig-out/bin/thottam ~/.local/bin/
\end{Verbatim}

\begin{note}[Build modes]
The default build uses ReleaseSafe (fast execution with bounds checking). For maximum performance, use \texttt{zig build -Doptimize=ReleaseFast}. Avoid Debug mode for normal use---it is roughly 500x slower on allocation-heavy workloads.
\end{note}

\section{Verifying Your Installation}

Check that Kaappi is installed correctly:

\begin{Verbatim}[fontsize=\small]
kaappi --version
\end{Verbatim}

You should see output like \texttt{Kaappi Scheme v\kaappiversion}. Next, test that evaluation works:

\begin{Verbatim}[fontsize=\small]
echo '(+ 1 2)' | kaappi
\end{Verbatim}

This should print \texttt{3}. If both commands succeed, your installation is ready.

When something is not working---or you simply want a thorough once-over---run the built-in self-check:

\begin{Verbatim}[fontsize=\small]
kaappi doctor
\end{Verbatim}

\texttt{doctor}\index{kaappi doctor}\index{doctor} inspects six areas (the binary, the library search path, the package manager, the native-compiler toolchain, the REPL, and the FFI) and prints \texttt{PASS}, \texttt{WARN}, or \texttt{FAIL} for each check, with a suggested fix under every failure:

\begin{Verbatim}[fontsize=\small, commandchars=\\\{\}]
binary
  PASS  version: v\kaappiversion
  PASS  target: aarch64-macos-none
  PASS  on-PATH: /Users/ada/.local/bin/kaappi

library
  PASS  ~/.kaappi/lib: /Users/ada/.kaappi/lib
  ...
\end{Verbatim}

Keep \texttt{kaappi doctor} in mind for later chapters too: it is the fastest way to diagnose a missing library path or an FFI toolchain problem.

\begin{note}[No install yet?]
If you want to follow along before installing, Kaappi provides a browser-based playground\index{playground} at \texttt{https://kaappi-lang.org/\allowbreak playground/} that runs the same interpreter via WebAssembly. It is perfect for trying examples, though you will eventually want a local install for file I/O and the full feature set.
\end{note}

\section{The REPL}
\index{REPL}

The REPL (Read-Eval-Print Loop) is where you will spend most of your time while learning. Launch it by running \texttt{kaappi} with no arguments:

\begin{Verbatim}[fontsize=\small]
kaappi
\end{Verbatim}

You will see a welcome banner and a prompt:

\begin{lstlisting}[style=repl]
Kaappi Scheme v(*@\kaappiversion@*)
Type ,help for commands, ,quit to exit.

kaappi> (+ 2 3)
5
kaappi> (* 6 7)
42
\end{lstlisting}

\begin{note}[Prompt convention]
The actual prompt is \texttt{kaappi>}. To keep examples compact, the rest of this book typesets it as a plain \texttt{>}.
\end{note}

\subsection{Multi-Line Input}

The REPL tracks open parentheses. If your expression is incomplete, it shows a \texttt{...} continuation prompt and waits for you to finish:

\begin{lstlisting}[style=repl]
> (define (square x)
...   (* x x))
> (square 7)
49
\end{lstlisting}

\subsection{Line Editing}

The REPL supports standard editing keys:

\begin{center}
\begin{tabular}{ll}
\textbf{Key} & \textbf{Action} \\
\hline
Left / Right & Move cursor \\
Up / Down & Navigate history \\
Ctrl-A & Move to start of line \\
Ctrl-E & Move to end of line \\
Ctrl-K & Delete to end of line \\
Ctrl-C & Cancel current line \\
Tab & Auto-complete symbol names \\
Ctrl-D & Exit (on empty line) \\
\end{tabular}
\end{center}

\subsection{Tab Completion}

Press Tab to complete symbol names. If you type \texttt{string-} and press Tab, the REPL shows all matching procedures:

\begin{lstlisting}[style=repl]
> string-
string-append  string-copy  string-length  string-ref  ...
\end{lstlisting}

Completion works for built-in procedures, your own definitions, and imported library exports.

\subsection{Syntax Highlighting}

As you type, the REPL colors different elements---keywords, strings, numbers, comments---and highlights the parenthesis that matches the one under your cursor. This makes it easier to spot typos and unbalanced parentheses.

The colors are configurable through \texttt{\textasciitilde/.kaappi/config}\index{config file}, a plain \texttt{key: value} file. \texttt{repl.theme} selects the \texttt{dark} (default) or \texttt{light} preset, \texttt{repl.color.\emph{token}} overrides individual token colors, and \texttt{repl.prompt} replaces the prompt text:

\begin{Verbatim}[fontsize=\small]
# ~/.kaappi/config
repl.theme: light
repl.color.string: cyan
repl.prompt: scheme>
\end{Verbatim}

Setting the \texttt{NO\_COLOR}\index{NO\_COLOR} environment variable disables all coloring, and \texttt{repl.highlight: false} turns highlighting off from the config file.

\subsection{The \texttt{\_} Variable}

The special variable \texttt{\_} always holds the result of the last evaluated expression:

\begin{lstlisting}[style=repl]
> (+ 10 20)
30
> (* _ 2)
60
\end{lstlisting}

This is useful for building on previous results without re-typing them.

\subsection{History}

Command history is saved to \texttt{\textasciitilde/.kaappi/history} and persists across sessions. Use the Up and Down arrow keys to navigate previous inputs.

\subsection{Exiting}

Type \texttt{,quit} or press Ctrl-D on an empty line to exit the REPL.

\section{Running Script Files}

While the REPL is great for exploration, real programs live in files. Create a file called \texttt{hello.scm}:

\begin{lstlisting}
(import (scheme base) (scheme write))

(display "Hello, world!")
(newline)
\end{lstlisting}

Run it from the command line:

\begin{Verbatim}[fontsize=\small]
kaappi hello.scm
\end{Verbatim}

Output:

\begin{codeoutput}
Hello, world!
\end{codeoutput}

\begin{note}[Why the import line?]
Kaappi pre-loads all of its built-in procedures, so this script would run even without the \texttt{import}\index{import} line. Writing it anyway is a good habit: it documents which libraries your code depends on, and other R7RS implementations require it---they raise an ``unbound variable'' error for anything not imported. Script files in this book include \texttt{import} lines for that reason. Chapter~\ref{ch:libraries} covers the library system in detail.
\end{note}

\section{Editor Support}
\index{editor support}\index{LSP}

You can write Scheme in any text editor, but language server support makes the experience significantly better.

\subsection{VS Code}

Install the \texttt{kaappi} extension from the VS Code marketplace. It provides syntax highlighting, bracket matching, and full LSP integration (go-to-definition, find-references, document symbols) via the bundled \texttt{kaappi-lsp}\index{kaappi-lsp} server.

\subsection{Neovim, Emacs, and Helix}

Configure your editor's LSP client to use \texttt{kaappi-lsp} as the language server for \texttt{.scm} and \texttt{.sld} files. Note that the install script does not include \texttt{kaappi-lsp}: build it from source (\texttt{zig build} produces it at \texttt{zig-out/bin/kaappi-lsp} alongside the interpreter). The VS Code extension needs no extra step because it bundles its own copy.

\subsection{Shell Completions}

Enable tab completion for the \texttt{kaappi} command itself:

\begin{Verbatim}[fontsize=\small]
# For zsh (add to ~/.zshrc)
eval "$(kaappi --completions zsh)"

# For bash (add to ~/.bashrc)
eval "$(kaappi --completions bash)"

# For fish (add to config.fish)
kaappi --completions fish | source
\end{Verbatim}

\begin{note}[Editor setup is optional]
For the purposes of this book, the REPL and a plain text editor are sufficient. Do not let editor configuration slow you down---you can always set up LSP later.
\end{note}

\section{The Package Manager}
\index{thottam}\index{package manager}

Kaappi ships with \texttt{thottam}, a package manager for installing third-party libraries. You will not need it for the first several chapters, but here is a quick preview of how it works.

Install a library:

\begin{Verbatim}[fontsize=\small]
thottam install kaappi-json
\end{Verbatim}

This downloads the \texttt{kaappi-json} library and its dependencies to \texttt{\textasciitilde/.kaappi/lib/}. Once installed, you can import it in any Scheme file:

\begin{lstlisting}
(import (kaappi json))

(define data (json-read (open-input-file "config.json")))
\end{lstlisting}

List installed packages:

\begin{Verbatim}[fontsize=\small]
thottam list
\end{Verbatim}

The ecosystem includes libraries for HTTP, web frameworks, PostgreSQL, SQLite, Redis, JSON, TOML, YAML, CSV, email, cryptography, logging, testing, and CLI tools. We will use several of these in later chapters; Appendix~\ref{app:ecosystem} lists them with brief descriptions.

\section*{Summary}

\begin{itemize}
    \item Install Kaappi with the one-line script, a manual binary download, or a source build, and confirm it with \texttt{kaappi --version}.
    \item The REPL provides multi-line input, line editing, tab completion, syntax highlighting, history, and the \texttt{\_} result variable.
    \item Run programs from files with \texttt{kaappi file.scm}; editors integrate through Kaappi's LSP server.
    \item The \texttt{thottam} package manager installs third-party libraries into \texttt{\textasciitilde/.kaappi/lib/}, which is always on the search path.
\end{itemize}

\section*{Exercises}

\begin{exercise}[Explore the REPL]
Launch the Kaappi REPL and experiment with its interactive features. Try each of the following:

\begin{enumerate}
    \item Type \code{string-} and press Tab to see all string procedures. Pick one you have not used before and call it with an argument.
    \item Define a variable, press Up to recall the definition, then edit it and press Enter to redefine it.
    \item Enter an incomplete expression (e.g., type \code{(define (greet name)} and press Enter) to see the continuation prompt. Complete the function body on the next line.
\end{enumerate}
\end{exercise}

\begin{exercise}[Your First Script File]
Create a file called \texttt{greet.scm} with the following content, then run it with \code{kaappi greet.scm}:

\begin{lstlisting}
(import (scheme base) (scheme write))

(define (greet name)
  (display "Hello, ")
  (display name)
  (display "!")
  (newline))

(greet "World")
(greet "Kaappi")
\end{lstlisting}

Verify that it prints two greeting lines:

\begin{codeoutput}
Hello, World!
Hello, Kaappi!
\end{codeoutput}

Then modify it to greet your own name.
\end{exercise}

\begin{exercise}[REPL Comma-Commands]
The REPL provides special commands prefixed with a comma. Type \code{,help} at the REPL prompt to see the full list. Try each of the following and observe what happens:

\begin{lstlisting}[style=repl]
> ,help
> ,env
> (define x 42)
> ,env
> ,quit
\end{lstlisting}

What does \code{,env} show before and after you define \code{x}? What other comma-commands look useful for debugging?
\end{exercise}

Appendix~\ref{app:cli} is a complete reference for the command-line interface---commands, options, environment variables, and REPL comma-commands---when you need to look something up. In the next chapter, we will start writing Scheme in earnest.

\chapterend
