\chapter{Functions}
\index{functions}

Functions are the primary organizational unit in Scheme. There are no classes, no methods, no modules in the traditional sense---just functions. This chapter covers defining functions, recursion, closures, tail-call optimization, variadic arguments, and returning multiple values.

\section{Defining Functions}
\index{define}

You have seen the shorthand form for defining functions:

\begin{lstlisting}[style=repl]
> (define (square x) (* x x))
> (square 7)
49
\end{lstlisting}

This is syntactic sugar for binding a name to a \texttt{lambda}\index{lambda} expression:

\begin{lstlisting}[style=repl]
> (define square (lambda (x) (* x x)))
> (square 7)
49
\end{lstlisting}

Both forms are equivalent. The shorthand \texttt{(define (name params) body)} is more common in practice, but understanding the \texttt{lambda} form is essential because anonymous functions appear everywhere in Scheme.

\subsection{Multiple Expressions in the Body}

Unlike Python's \texttt{lambda} (limited to one expression), a Scheme lambda body can contain multiple expressions. The value of the last expression is returned:

\begin{lstlisting}[style=repl]
> (define (greet name)
    (display "Hello, ")
    (display name)
    (newline)
    (string-append "Greeted " name))
> (greet "Alice")
Hello, Alice
"Greeted Alice"
\end{lstlisting}

The first three expressions produce side effects (printing). The last expression's value is the function's return value.

\subsection{Functions with Multiple Parameters}

\begin{lstlisting}[style=repl]
> (define (hypotenuse a b)
    (sqrt (+ (* a a) (* b b))))
> (hypotenuse 3 4)
5
\end{lstlisting}

\section{Anonymous Functions}
\index{lambda}\index{anonymous functions}

You do not need to name every function. \texttt{lambda} creates a function value directly:

\begin{lstlisting}[style=repl]
> ((lambda (x) (* x x)) 5)
25
> (map (lambda (x) (+ x 10)) '(1 2 3 4))
(11 12 13 14)
\end{lstlisting}

Anonymous functions are particularly useful when passing short operations to higher-order functions like \texttt{map} and \texttt{filter}. You will use them constantly.

\begin{note}[Coming from Python]
\texttt{(lambda (x) (* x x))} is Scheme's equivalent of Python's \texttt{lambda x: x * x}. The difference is that Scheme's version can contain multiple expressions and arbitrary complexity---there is no need for a separate \texttt{def} just because the function has more than one line.
\end{note}

\section{Recursion}
\index{recursion}

Scheme has no \texttt{for} or \texttt{while} loops. Repetition is expressed through recursion, and the language is designed to make this efficient and natural.

\subsection{Simple Recursion}

The classic factorial:

\begin{lstlisting}[style=repl]
> (define (factorial n)
    (if (= n 0)
        1
        (* n (factorial (- n 1)))))
> (factorial 5)
120
> (factorial 20)
2432902008176640000
\end{lstlisting}

The structure is always the same: a base case that returns a known value, and a recursive case that reduces the problem.

Here is a function that computes the nth Fibonacci number:

\begin{lstlisting}[style=repl]
> (define (fib n)
    (cond
      ((= n 0) 0)
      ((= n 1) 1)
      (else (+ (fib (- n 1))
               (fib (- n 2))))))
> (fib 10)
55
\end{lstlisting}

\begin{warning}[Exponential recursion]
This naive Fibonacci is O($2^n$)---it recomputes the same subproblems repeatedly. For large \texttt{n}, use the tail-recursive version shown later in this chapter.
\end{warning}

\subsection{Recursion on Lists}

Most list processing follows this pattern: check for empty (base case), process the first element, recurse on the rest:

\begin{lstlisting}[style=repl]
> (define (my-map f lst)
    (if (null? lst)
        '()
        (cons (f (car lst))
              (my-map f (cdr lst)))))
> (my-map square '(1 2 3 4 5))
(1 4 9 16 25)
\end{lstlisting}

\section{Tail Calls and Tail Recursion}
\index{tail call optimization}\index{TCO}

When a recursive call is the \emph{last thing} a function does---the final expression with nothing left to do after it returns---it is in \emph{tail position}. Kaappi (and all conforming Schemes) guarantee that tail calls reuse the current stack frame instead of allocating a new one.

This means tail-recursive functions use constant stack space, just like a loop.

\subsection{Non-Tail vs.\ Tail Recursive}

The factorial above is \emph{not} tail-recursive because after the recursive call, it still needs to multiply:

\begin{lstlisting}
;; NOT tail-recursive: (* n ...) wraps the recursive call
(define (factorial n)
  (if (= n 0) 1
      (* n (factorial (- n 1)))))
\end{lstlisting}

Here is the tail-recursive version using an accumulator:

\begin{lstlisting}[style=repl]
> (define (factorial n)
    (define (loop i acc)
      (if (= i 0)
          acc
          (loop (- i 1) (* acc i))))
    (loop n 1))
> (factorial 100)
93326215443944152681699238856266700490715968264381621468592963895217599993229915608941463976156518286253697920827223758251185210916864000000000000000000000000
\end{lstlisting}

The call to \texttt{loop} is the last expression---nothing wraps it---so it runs without growing the stack, even for \texttt{(factorial 1000000)}.

\subsection{Named \texttt{let} for Loops}
\index{named let}

Named \texttt{let} is the idiomatic way to write a tail-recursive loop. It defines a local function and calls it immediately:

\begin{lstlisting}[style=repl]
> (let loop ((i 1))
    (when (<= i 5)
      (display i)
      (display " ")
      (loop (+ i 1))))
1 2 3 4 5
\end{lstlisting}

\texttt{loop} is a name you choose---it is bound to a function with parameter \texttt{i}. Calling \texttt{(loop (+ i 1))} recurses with the next value. Because the call is in tail position, it runs in constant space.

Here is a tail-recursive Fibonacci using named \texttt{let}:

\begin{lstlisting}[style=repl]
> (define (fib n)
    (let loop ((i n) (a 0) (b 1))
      (if (= i 0)
          a
          (loop (- i 1) b (+ a b)))))
> (fib 50)
12586269025
\end{lstlisting}

This version is O(n), computing \texttt{(fib 50)} instantly.

\begin{note}[Coming from Python]
Named \texttt{let} is Scheme's equivalent of a \texttt{while} loop with loop variables. The ``loop variables'' are the parameters, and each ``iteration'' is a tail call with updated values. Once you internalize this pattern, you will find it as natural as \texttt{for i in range(...)}.
\end{note}

\section{Closures}
\index{closures}

A \emph{closure} is a function that remembers the environment where it was created. Variables from the enclosing scope remain accessible even after that scope has exited.

\begin{lstlisting}[style=repl]
> (define (make-adder n)
    (lambda (x) (+ x n)))
> (define add5 (make-adder 5))
> (define add10 (make-adder 10))
> (add5 3)
8
> (add10 3)
13
\end{lstlisting}

Each call to \texttt{make-adder} creates a new closure that captures its own copy of \texttt{n}. The two closures \texttt{add5} and \texttt{add10} are independent.

\subsection{Closures with Mutable State}

Closures can capture mutable variables, creating private state:

\begin{lstlisting}[style=repl]
> (define (make-counter)
    (let ((n 0))
      (lambda ()
        (set! n (+ n 1))
        n)))
> (define c1 (make-counter))
> (define c2 (make-counter))
> (c1)
1
> (c1)
2
> (c1)
3
> (c2)
1
\end{lstlisting}

Each counter has its own private \texttt{n}. This is how you encapsulate state in Scheme without classes.

\begin{note}[Coming from Python]
This is the same concept as Python closures. The difference is that Scheme uses closures as a primary organizational tool (where Python might reach for a class), and \texttt{set!}\index{set!} makes the captured variable mutable.
\end{note}

\section{Variadic Functions}
\index{variadic functions}\index{rest arguments}

Functions can accept a variable number of arguments using \emph{dot notation} in the parameter list:

\begin{lstlisting}[style=repl]
> (define (sum . numbers)
    (let loop ((lst numbers) (acc 0))
      (if (null? lst)
          acc
          (loop (cdr lst) (+ acc (car lst))))))
> (sum 1 2 3)
6
> (sum 1 2 3 4 5)
15
\end{lstlisting}

The dot before \texttt{numbers} means ``collect all remaining arguments into a list.'' You can also have required parameters before the rest:

\begin{lstlisting}[style=repl]
> (define (format-message prefix . parts)
    (apply string-append prefix ": " parts))
> (format-message "ERROR" "file not found")
"ERROR: file not found"
\end{lstlisting}

\section{\texttt{apply}}
\index{apply}

\texttt{apply} calls a function with arguments from a list:

\begin{lstlisting}[style=repl]
> (apply + '(1 2 3 4 5))
15
> (apply max '(3 7 2 9 4))
9
> (apply string-append '("hello" " " "world"))
"hello world"
\end{lstlisting}

This is essential when you have a list of arguments and need to pass them to a function that expects individual parameters.

\section{Multiple Return Values}
\index{values}\index{multiple values}

Scheme functions can return multiple values using \texttt{values}:

\begin{lstlisting}[style=repl]
> (define (divide a b)
    (values (quotient a b) (remainder a b)))
> (divide 17 5)
3
2
\end{lstlisting}

To receive multiple values, use \texttt{call-with-values} or the more convenient \texttt{let-values}:

\begin{lstlisting}[style=repl]
> (let-values (((q r) (divide 17 5)))
    (display "Quotient: ") (display q) (newline)
    (display "Remainder: ") (display r) (newline))
Quotient: 3
Remainder: 2
\end{lstlisting}

\begin{note}[Coming from Python]
This is similar to Python's tuple unpacking: \texttt{q, r = divmod(17, 5)}. Scheme's version is slightly more verbose but achieves the same result.
\end{note}

\section{Function Composition Patterns}

A few patterns come up frequently when working with functions.

\subsection{Functions that Return Functions}

\begin{lstlisting}[style=repl]
> (define (compose f g)
    (lambda (x) (f (g x))))
> (define inc-then-square (compose square (lambda (x) (+ x 1))))
> (inc-then-square 4)
25
\end{lstlisting}

\subsection{Partial Application}

\begin{lstlisting}[style=repl]
> (define (partial f . fixed-args)
    (lambda rest-args
      (apply f (append fixed-args rest-args))))
> (define double (partial * 2))
> (double 5)
10
> (define add1 (partial + 1))
> (add1 99)
100
\end{lstlisting}

These patterns---composition, partial application, and higher-order functions in general---are the tools Scheme gives you in place of classes and inheritance. They are simpler, more composable, and often more powerful.

\sectionbreak

You now understand how to define functions, write recursive algorithms with tail-call guarantees, create closures with private state, handle variable arguments, and compose functions. These are the core skills of Scheme programming. The next chapter covers control flow in depth.
