\chapter{Higher-Order Functions}
\index{higher-order functions}
\label{ch:hof}

A \emph{higher-order function} is a function that takes another function as an argument or returns a function as its result. You have already seen examples in earlier chapters---\texttt{map}, named \texttt{let} with a procedure argument, and closures. This chapter explores the standard higher-order functions in depth and shows how they replace loops in idiomatic Scheme.

\begin{note}[SRFI-1 imports]
Several procedures in this chapter---\texttt{filter}, \texttt{remove}, \texttt{partition}, \texttt{any}, \texttt{every}, \texttt{fold}, and \texttt{fold-right}---come from SRFI-1\index{SRFI-1} (the extended list library), not from \texttt{(scheme base)}. In script files, add \texttt{(import (srfi 1))} at the top. The REPL makes them available automatically.
\end{note}

\section{\texttt{map}: Transform Every Element}
\index{map}

\texttt{map} applies a function to every element of a list, returning a new list of the results:

\begin{lstlisting}[style=repl]
> (define (square x) (* x x))
> (map square '(1 2 3 4 5))
(1 4 9 16 25)
> (map string-length '("cat" "elephant" "ox"))
(3 8 2)
> (map not '(#t #f #t #f))
(#f #t #f #t)
\end{lstlisting}

With a lambda for custom transformations:

\begin{lstlisting}[style=repl]
> (map (lambda (x) (+ x 10)) '(1 2 3))
(11 12 13)
> (map (lambda (s) (string-append "Hello, " s))
       '("Alice" "Bob" "Carol"))
("Hello, Alice" "Hello, Bob" "Hello, Carol")
\end{lstlisting}

\subsection{Multi-List \texttt{map}}

\texttt{map} can take multiple lists, applying the function to corresponding elements:

\begin{lstlisting}[style=repl]
> (map + '(1 2 3) '(10 20 30))
(11 22 33)
> (map * '(2 3 4) '(10 10 10))
(20 30 40)
> (map list '(a b c) '(1 2 3))
((a 1) (b 2) (c 3))
> (map + '(1 2 3) '(10 20))
(11 22)
\end{lstlisting}

The function receives one argument from each list. If the lists have different lengths, \texttt{map} stops at the end of the shortest one, as the last example shows.

\begin{note}[Coming from Python]
\texttt{(map f lst)} is equivalent to Python's \texttt{[f(x) for x in lst]} or \texttt{list(map(f, lst))}. Multi-list map is like \texttt{[f(a, b) for a, b in zip(lst1, lst2)]}.
\end{note}

\section{\texttt{for-each}: Side Effects Only}
\index{for-each}

\texttt{for-each} is like \texttt{map} but discards the results. Use it when you want side effects (printing, writing to a file) without building a list:

\begin{lstlisting}[style=repl]
> (for-each (lambda (name)
              (display "Hello, ")
              (display name)
              (newline))
            '("Alice" "Bob" "Carol"))
Hello, Alice
Hello, Bob
Hello, Carol
\end{lstlisting}

\texttt{for-each} guarantees left-to-right evaluation order. \texttt{map} does not make this guarantee (though most implementations process left to right).

\begin{note}[Coming from Python]
\texttt{for-each} replaces a Python \texttt{for} loop where you do not need the return value:
\begin{Verbatim}[fontsize=\small]
for name in ["Alice", "Bob", "Carol"]:
    print(f"Hello, {name}")
\end{Verbatim}
\end{note}

\section{\texttt{filter}: Select Elements}
\index{filter}

\texttt{filter} keeps only the elements for which a predicate returns true:

\begin{lstlisting}[style=repl]
> (filter even? '(1 2 3 4 5 6 7 8))
(2 4 6 8)
> (filter positive? '(-3 -1 0 2 5 -4 7))
(2 5 7)
> (filter (lambda (s) (> (string-length s) 3))
          '("hi" "hello" "hey" "good morning"))
("hello" "good morning")
\end{lstlisting}

\subsection{\texttt{remove}: The Complement of \texttt{filter}}
\index{remove}

\texttt{remove} (also from SRFI-1) keeps elements that do \emph{not} satisfy the predicate:

\begin{lstlisting}[style=repl]
> (remove even? '(1 2 3 4 5 6))
(1 3 5)
> (remove null? '((1 2) () (3) () (4 5)))
((1 2) (3) (4 5))
\end{lstlisting}

\subsection{\texttt{partition}: Split into Two Lists}
\index{partition}

\texttt{partition} returns two values: elements that pass and elements that fail:

\begin{lstlisting}[style=repl]
> (let-values (((yes no) (partition even? '(1 2 3 4 5 6))))
    (list yes no))
((2 4 6) (1 3 5))
\end{lstlisting}

\section{\texttt{any} and \texttt{every}: Testing Across a List}
\index{any}\index{every}

\texttt{any} returns a true value if the predicate holds for at least one element; \texttt{every} checks that it holds for all of them:

\begin{lstlisting}[style=repl]
> (any even? '(1 3 5 6))
#t
> (any even? '(1 3 5))
#f
> (every positive? '(1 2 3))
#t
> (every positive? '(1 -2 3))
#f
\end{lstlisting}

Both short-circuit: \texttt{any} stops at the first hit, \texttt{every} at the first miss. And like \texttt{and} and \texttt{or}, \texttt{any} returns the predicate's actual result rather than a bare \texttt{\#t}---useful when the test computes something worth keeping:

\begin{lstlisting}[style=repl]
> (any string->number '("a" "b" "42"))
42
\end{lstlisting}

\begin{note}[Coming from Python]
\texttt{any} and \texttt{every} correspond to Python's \texttt{any()} and \texttt{all()} over a generator expression: \texttt{(any even? lst)} is \texttt{any(x \% 2 == 0 for x in lst)}.
\end{note}

\section{\texttt{fold}: Reduce a List to a Single Value}
\index{fold}

\texttt{fold} (also called ``reduce'' in other languages) combines all elements of a list into one value using a binary function and a starting value:

\begin{lstlisting}[style=repl]
> (fold + 0 '(1 2 3 4 5))
15
> (fold * 1 '(1 2 3 4 5))
120
> (fold max 0 '(3 7 2 9 4))
9
> (fold string-append "" '("a" "b" "c"))
"cba"
\end{lstlisting}

Why did \texttt{string-append} produce \texttt{"cba"} instead of \texttt{"abc"}? Because \texttt{fold} passes each element as the \emph{first} argument and the accumulator as the \emph{second}:

\begin{lstlisting}
;; (fold string-append "" '("a" "b" "c"))
;; Step 1: (string-append "a" "")  = "a"
;; Step 2: (string-append "b" "a") = "ba"
;; Step 3: (string-append "c" "ba") = "cba"
\end{lstlisting}

The form is \texttt{(fold f init lst)}. The function \texttt{f} takes two arguments: the current element and the accumulator. Processing is left to right, but the accumulator is the \emph{second} argument:

\begin{lstlisting}[style=repl]
> (fold cons '() '(1 2 3 4 5))
(5 4 3 2 1)
\end{lstlisting}

\subsection{Understanding Fold Step by Step}

\texttt{(fold + 0 '(1 2 3))} evaluates as:

\begin{lstlisting}
;; Start with init = 0
;; Step 1: (+ 1 0) = 1
;; Step 2: (+ 2 1) = 3
;; Step 3: (+ 3 3) = 6
;; Result: 6
\end{lstlisting}

\begin{lstlisting}[style=repl]
> (fold + 0 '(1 2 3))
6
\end{lstlisting}

\subsection{\texttt{fold-right}: Right-to-Left Reduction}
\index{fold-right}

\texttt{fold-right} processes elements from right to left:

\begin{lstlisting}[style=repl]
> (fold-right cons '() '(1 2 3 4 5))
(1 2 3 4 5)
> (fold-right string-append "" '("a" "b" "c"))
"abc"
\end{lstlisting}

\texttt{fold-right} with \texttt{cons} copies a list (preserving order), while \texttt{fold} with \texttt{cons} reverses it.

\subsection{Expressing Other Operations as Fold}

Many list operations are special cases of fold:

\begin{lstlisting}[style=repl]
> ;; sum
> (fold + 0 '(1 2 3 4 5))
15
> ;; length
> (fold (lambda (x acc) (+ acc 1)) 0 '(a b c d))
4
> ;; reverse
> (fold cons '() '(1 2 3))
(3 2 1)
> ;; map (via fold-right)
> (fold-right (lambda (x acc) (cons (* x x) acc))
              '() '(1 2 3 4))
(1 4 9 16)
> ;; filter (via fold-right)
> (fold-right (lambda (x acc)
               (if (even? x) (cons x acc) acc))
             '() '(1 2 3 4 5 6))
(2 4 6)
\end{lstlisting}

\begin{note}[Coming from Python]
\texttt{fold} is Python's \texttt{functools.reduce}. \texttt{(fold + 0 '(1 2 3 4 5))} is \texttt{reduce(lambda acc, x: acc + x, [1,2,3,4,5], 0)}. But watch the argument order: it is the \emph{reverse} of Python's. \texttt{reduce} passes the accumulator first (\texttt{acc, x}); \texttt{fold} passes the element first (\texttt{x acc}). Swapping the two is the classic mistake when porting \texttt{reduce} code---if your fold returns a reversed list or a baffling type error, check the argument order first.
\end{note}

\section{\texttt{apply}: Spreading a List as Arguments}
\index{apply}

\texttt{apply}, introduced in Chapter~\ref{ch:functions}, calls a function with the elements of a list as individual arguments---\code{(apply + '(1 2 3 4 5))} is \code{15}. One detail that chapter used without comment: you can provide fixed arguments before the list:

\begin{lstlisting}[style=repl]
> (apply + 100 '(1 2 3))
106
> (apply list 'a 'b '(c d e))
(a b c d e)
\end{lstlisting}

\section{Composing Higher-Order Functions}

The real power emerges when you combine these operations into pipelines:

\begin{lstlisting}[style=repl]
> ;; Sum of squares of even numbers
> (fold + 0
    (map square
      (filter even? '(1 2 3 4 5 6 7 8 9 10))))
220
\end{lstlisting}

Read from the inside out: filter the even numbers, square each one, then sum them. In Python this would be:

\begin{Verbatim}[fontsize=\small]
sum(x**2 for x in range(1, 11) if x % 2 == 0)
\end{Verbatim}

Another example---find the longest string in a list:

\begin{lstlisting}[style=repl]
> (fold (lambda (s longest)
          (if (> (string-length s) (string-length longest))
              s
              longest))
        ""
        '("cat" "elephant" "dog" "hippopotamus"))
"hippopotamus"
\end{lstlisting}

\section{Writing Your Own Higher-Order Functions}

You are not limited to the standard ones. Here are a few useful patterns:

\subsection{Applying a Function \texttt{n} Times}

\begin{lstlisting}[style=repl]
> (define (iterate f n x)
    (if (= n 0) x
        (iterate f (- n 1) (f x))))
> (iterate (lambda (x) (* x 2)) 5 1)
32
> (iterate cdr 3 '(a b c d e))
(d e)
\end{lstlisting}

\subsection{Finding the First Match}

SRFI-1 already provides this one as \texttt{find}\index{find}---but it is worth seeing what is inside:

\begin{lstlisting}[style=repl]
> (define (my-find pred lst)
    (cond
      ((null? lst) #f)
      ((pred (car lst)) (car lst))
      (else (my-find pred (cdr lst)))))
> (my-find even? '(1 3 5 4 7))
4
> (my-find negative? '(1 2 3))
#f
\end{lstlisting}

\subsection{Counting Matches}

Also built in as SRFI-1's \texttt{count}\index{count}; here it is as a fold:

\begin{lstlisting}[style=repl]
> (define (my-count pred lst)
    (fold (lambda (x acc)
            (if (pred x) (+ acc 1) acc))
          0 lst))
> (my-count even? '(1 2 3 4 5 6))
3
> (my-count char-alphabetic?
            (string->list "hello, world!"))
10
\end{lstlisting}

\subsection{Zipping Two Lists}

\begin{lstlisting}[style=repl]
> (define (zip lst1 lst2)
    (map list lst1 lst2))
> (zip '(a b c) '(1 2 3))
((a 1) (b 2) (c 3))
\end{lstlisting}

\subsection{Building \texttt{filter} from Scratch}

The standard higher-order functions are ordinary Scheme code, not magic. Here is \texttt{filter}, written with the recursive list pattern from Chapter~\ref{ch:functions}:

\begin{lstlisting}
(define (my-filter pred lst)
  (cond
    ((null? lst) '())
    ((pred (car lst))
     (cons (car lst) (my-filter pred (cdr lst))))
    (else
     (my-filter pred (cdr lst)))))
\end{lstlisting}

\begin{lstlisting}[style=repl]
> (my-filter even? '(1 2 3 4 5 6))
(2 4 6)
> (my-filter string? '(1 "two" 3 "four"))
("two" "four")
\end{lstlisting}

Writing \texttt{map}, \texttt{filter}, or \texttt{fold} by hand once is worth the effort: afterward you know exactly what they do, what they cost, and how to build variants when the standard ones do not quite fit.

\section{When Not to Use Higher-Order Functions}

Higher-order functions are elegant, but they are not always the best choice:

\begin{itemize}
    \item \textbf{Complex state:} When the loop body needs to maintain multiple interacting pieces of state, named \texttt{let} with explicit accumulators is often clearer than contorting a fold.
    \item \textbf{Early termination:} \texttt{map}, \texttt{filter}, and \texttt{fold} always process the entire list. If you want to stop early, use \texttt{find}, \texttt{any}, or \texttt{every}---they return as soon as they have an answer---or a recursive function.
    \item \textbf{Multiple outputs:} If you need both the filtered elements and the rejected ones, use \texttt{partition} rather than calling \texttt{filter} twice.
\end{itemize}

\section*{Summary}

\begin{itemize}
    \item \texttt{map} transforms every element of a list, while \texttt{for-each} applies a procedure for its side effects.
    \item \texttt{filter}, \texttt{remove}, and \texttt{partition} select elements by a predicate, while \texttt{any} and \texttt{every} test whether it holds somewhere or everywhere.
    \item \texttt{fold} reduces a list to a single value and can express \texttt{map}, \texttt{filter}, \texttt{length}, and more.
    \item \texttt{apply} spreads a list as arguments, and higher-order functions compose into data-processing pipelines.
    \item Higher-order functions usually replace explicit recursion, though sometimes plain recursion is clearer.
\end{itemize}

\section*{Exercises}

\begin{exercise}[Extracting Data with \code{map} and \code{filter}]
Given this list of association lists representing people:

\begin{lstlisting}
(define people
  '(((name . "Alice") (age . 30) (city . "Paris"))
    ((name . "Bob") (age . 17) (city . "London"))
    ((name . "Carol") (age . 25) (city . "Paris"))
    ((name . "Dave") (age . 15) (city . "Berlin"))))
\end{lstlisting}

Write expressions (using \code{map}, \code{filter}, and \code{assq}/\code{assoc}) that:
\begin{enumerate}
\item Extract a list of all names: \code{("Alice" "Bob" "Carol" "Dave")}
\item Find the names of people aged 18 or older: \code{("Alice" "Carol")}
\item Find the names of the people who live in Paris: \code{("Alice" "Carol")}
\end{enumerate}
\end{exercise}

\begin{exercise}[Fold Implementations]
Implement each of these using \code{fold} (not explicit recursion):

\begin{enumerate}
\item \code{my-length}: returns the number of elements in a list
\item \code{my-reverse}: reverses a list
\end{enumerate}

\begin{lstlisting}[style=repl]
> (my-length '(a b c d e))
5
> (my-reverse '(1 2 3 4 5))
(5 4 3 2 1)
\end{lstlisting}

Hint: for \code{my-length}, the accumulator is a counter. For \code{my-reverse}, the accumulator is a list built with \code{cons}.
\end{exercise}

\begin{exercise}[Word Processing Pipeline]
Write a function \code{process-words} that takes a list of strings and:
\begin{enumerate}
\item Filters out words shorter than 4 characters
\item Sorts the remaining words alphabetically (use \code{list-sort} from SRFI-132: \code{(import (srfi 132))}---sorting is covered properly in Chapter~\ref{ch:stdlib}; here you only need \code{(list-sort string<? lst)})
\item Capitalizes each word (upper-case the first letter, lower-case the rest)
\end{enumerate}

\begin{lstlisting}[style=repl]
> (process-words '("the" "quick" "brown" "fox" "jumps"
                   "over" "the" "lazy" "dog"))
("Brown" "Jumps" "Lazy" "Over" "Quick")
\end{lstlisting}

Build this as a pipeline: feed the output of \code{filter} into \code{list-sort}, then \code{map} the capitalization function.
\end{exercise}

\begin{exercise}[Zipping Lists]
Implement \code{zip} that pairs up corresponding elements from two lists. If the lists have different lengths, stop at the shorter one:

\begin{lstlisting}[style=repl]
> (zip '(a b c) '(1 2 3))
((a 1) (b 2) (c 3))
> (zip '(x y) '(10 20 30))
((x 10) (y 20))
\end{lstlisting}

The chapter showed a one-liner using multi-list \code{map}, which already stops at the shorter list. Write the recursive version so it behaves the same way---what are the two base cases?
\end{exercise}

\begin{exercise}[\code{take} and \code{drop}]
Implement \code{take} and \code{drop}, which return the first $n$ elements or skip the first $n$ elements of a list. Use the SRFI-1 argument order (list first, count second):

\begin{lstlisting}[style=repl]
> (take '(a b c d e) 3)
(a b c)
> (drop '(a b c d e) 3)
(d e)
> (take '(1 2 3) 0)
()
> (drop '(1 2 3) 10)
()
\end{lstlisting}

Use recursion or named \code{let}. Handle the case where $n$ exceeds the list length gracefully. Verify that \code{(append (take lst n) (drop lst n))} reconstructs the original list.

SRFI-1 provides built-in \code{take} and \code{drop}---the purpose here is to practice the recursive pattern.
\end{exercise}

In the next chapter, we cover input and output---reading from files, writing to the console, and working with ports.

\chapterend
