\chapter{Input and Output}
\index{input/output}\index{ports}
\label{ch:io}

Programs need to communicate with the outside world: reading user input, writing results to the screen, processing files, and formatting data. Scheme's I/O system is based on \emph{ports}---abstract objects that represent sources and destinations for data. This chapter covers console I/O, file operations, string ports, and the difference between \texttt{display} and \texttt{write}.

\section{Output: \texttt{display}, \texttt{write}, and \texttt{newline}}
\index{display}\index{write}\index{newline}

Scheme provides two primary output procedures, each serving a different purpose.

\subsection{\texttt{display}: Human-Readable Output}

\texttt{display} outputs a value in a form meant for humans. Strings appear without quotes, characters without the \texttt{\#\textbackslash} prefix:

\begin{lstlisting}[style=repl]
> (display "Hello, world!")
Hello, world!
> (display 42)
42
> (display #\a)
a
> (display '(1 2 3))
(1 2 3)
\end{lstlisting}

\texttt{display} does not add a newline. Use \texttt{newline} to move to the next line:

\begin{lstlisting}[style=repl]
> (begin (display "line one") (newline)
         (display "line two") (newline))
line one
line two
\end{lstlisting}

\subsection{\texttt{write}: Machine-Readable Output}

\texttt{write} outputs a value in a form that can be read back by \texttt{read}. Strings include quotes, characters include the \texttt{\#\textbackslash} prefix:

\begin{lstlisting}[style=repl]
> (write "hello")
"hello"
> (write #\a)
#\a
> (write '(1 "two" #t))
(1 "two" #t)
\end{lstlisting}

Use \texttt{write} when you want to serialize data that will later be read by a program. Use \texttt{display} when you want output meant for a person.

\begin{note}[Coming from Python]
\texttt{display} is like Python's \texttt{print(..., end="")} (without the trailing newline). \texttt{write} is like Python's \texttt{repr()}---it shows the value in a way you could paste back into code.
\end{note}

\subsection{Combining Output}

A common pattern for formatted output:

\begin{lstlisting}[style=repl]
> (define (show-result label value)
    (display label)
    (display ": ")
    (display value)
    (newline))
> (show-result "Total" 42)
Total: 42
> (show-result "Name" "Alice")
Name: Alice
\end{lstlisting}

\section{Input: \texttt{read-line} and \texttt{read}}
\index{read-line}\index{read}

\subsection{\texttt{read-line}: Reading Text}

\texttt{read-line} reads one line of text from standard input (up to the newline, which is consumed but not included):

\begin{lstlisting}[style=repl]
> (define (prompt msg)
    (display msg)
    (flush-output-port)
    (read-line))
> (define name (prompt "Enter your name: "))
Enter your name: Alice
> (string-append "Hello, " name "!")
"Hello, Alice!"
\end{lstlisting}

The call to \texttt{flush-output-port}\index{flush-output-port} forces the prompt out of the output buffer so it is visible before \texttt{read-line} blocks waiting for input. Output may be buffered until a newline is written or the buffer fills; flush whenever a partial line must appear immediately.

At end of file, \texttt{read-line} returns an \emph{eof object}\index{eof object}. You can test for it with \texttt{eof-object?}:

\begin{lstlisting}
(let loop ((line (read-line)))
  (unless (eof-object? line)
    (display line)
    (newline)
    (loop (read-line))))
\end{lstlisting}

\subsection{\texttt{read}: Reading Scheme Data}

\texttt{read} parses a complete Scheme datum from input---a number, string, list, symbol, or any other value the reader understands:

\begin{lstlisting}[style=repl]
> (define x (read))
(1 2 3)
> (car x)
1
> (length x)
3
\end{lstlisting}

\texttt{read} is the inverse of \texttt{write}: anything produced by \texttt{write} can be read back by \texttt{read}. This makes Scheme's S-expression syntax a built-in serialization format.

\begin{warning}[Security]
Be careful with \texttt{read} on untrusted input. Unlike Python's \texttt{eval}, it cannot execute code---it only builds data structures---but a hostile sender can still exhaust memory with a huge or deeply nested datum. For untrusted data, read a bounded amount with \texttt{read-line} and parse the string yourself.
\end{warning}

\section{Ports}
\index{ports}

A \emph{port} is an abstract source or destination for I/O. Every read operation reads from some port; every write operation writes to some port. When you call \texttt{display} or \texttt{read-line} without specifying a port, they use the \emph{current ports}\index{current-input-port}\index{current-output-port}\index{current-error-port}:

\begin{lstlisting}[style=repl]
> (current-input-port)
#<input-port stdin>
> (current-output-port)
#<output-port stdout>
> (current-error-port)
#<output-port stderr>
\end{lstlisting}

You can write to a specific port by passing it as an extra argument:

\begin{lstlisting}[style=repl]
> (display "error!\n" (current-error-port))
error!
\end{lstlisting}

\begin{note}[Current ports are parameters]
\texttt{current-input-port}, \texttt{current-output-port}, and \texttt{current-error-port} are ordinary parameter objects (Chapter~\ref{ch:bindings}). That means you can redirect all output within a computation using \texttt{parameterize}---no procedure needs to be told about the redirection explicitly.
\end{note}

\subsection{Port Predicates}
\index{port?}\index{input-port?}\index{output-port?}

\begin{lstlisting}[style=repl]
> (port? (current-input-port))
#t
> (input-port? (current-input-port))
#t
> (output-port? (current-output-port))
#t
\end{lstlisting}

\section{File I/O}
\index{file I/O}

\subsection{Opening and Closing Ports}
\index{open-input-file}\index{open-output-file}\index{close-input-port}\index{close-output-port}

\begin{lstlisting}
(define in (open-input-file "data.txt"))
;; ... read from in ...
(close-input-port in)

(define out (open-output-file "output.txt"))
;; ... write to out ...
(close-output-port out)
\end{lstlisting}

Two behaviors to know before you rely on these. Opening an input file that does not exist raises an error---Chapter~\ref{ch:errors} shows how to catch it with the \texttt{file-error?} predicate. And opening an output file that already exists \emph{truncates} it: the old contents are gone the moment the port opens.

\subsection{\texttt{call-with-port}: Safe Port Handling}
\index{call-with-port}

To ensure ports are always closed (even if an error occurs), use \texttt{call-with-port}:

\begin{lstlisting}
(call-with-port (open-input-file "data.txt")
  (lambda (port)
    (let loop ((line (read-line port)))
      (unless (eof-object? line)
        (display line)
        (newline)
        (loop (read-line port))))))
\end{lstlisting}

The port is automatically closed when the procedure returns.

This open-then-call combination is common enough that R7RS provides shorthands: \texttt{call-with-input-file}\index{call-with-input-file} and \texttt{call-with-output-file}\index{call-with-output-file} take the filename directly:

\begin{lstlisting}
(call-with-input-file "data.txt"
  (lambda (port)
    (read-line port)))
\end{lstlisting}

\subsection{\texttt{with-input-from-file} and \texttt{with-output-to-file}}
\index{with-input-from-file}\index{with-output-to-file}

These redirect the current input/output port for the duration of a thunk (a zero-argument procedure):

\begin{lstlisting}
;; Read all lines from a file
(with-input-from-file "data.txt"
  (lambda ()
    (let loop ((line (read-line)))
      (unless (eof-object? line)
        (display line)
        (newline)
        (loop (read-line))))))
\end{lstlisting}

\begin{lstlisting}
;; Write to a file
(with-output-to-file "output.txt"
  (lambda ()
    (display "Hello, file!")
    (newline)))
\end{lstlisting}

\begin{note}[Coming from Python]
\texttt{call-with-port} is similar to Python's \texttt{with open(...) as f:} context manager---it guarantees the file is closed when you are done.
\end{note}

\subsection{Reading an Entire File}

A utility that reads all lines into a list:

\begin{lstlisting}
(define (read-all-lines filename)
  (call-with-port (open-input-file filename)
    (lambda (port)
      (let loop ((line (read-line port))
                 (lines '()))
        (if (eof-object? line)
            (reverse lines)
            (loop (read-line port)
                  (cons line lines)))))))
\end{lstlisting}

\subsection{Writing to a File}

\begin{lstlisting}
(define (write-lines filename lines)
  (call-with-port (open-output-file filename)
    (lambda (port)
      (for-each (lambda (line)
                  (display line port)
                  (newline port))
                lines))))
\end{lstlisting}

\subsection{\texttt{file-exists?} and \texttt{delete-file}}
\index{file-exists?}\index{delete-file}

Two housekeeping procedures round out file handling. Here we use the \texttt{write-lines} helper from above to create a file, then remove it:

\begin{lstlisting}[style=repl]
> (write-lines "data.txt" '("some" "content"))
> (file-exists? "data.txt")
#t
> (delete-file "data.txt")
> (file-exists? "data.txt")
#f
\end{lstlisting}

Check \texttt{file-exists?} before opening a file for output when silently overwriting it would be destructive.

\section{String Ports}
\index{string ports}

String ports let you use I/O procedures on strings instead of files. They are useful for building strings incrementally or parsing string content.

\subsection{Output String Ports}
\index{open-output-string}\index{get-output-string}

\begin{lstlisting}[style=repl]
> (define port (open-output-string))
> (display "Hello, " port)
> (display "world!" port)
> (get-output-string port)
"Hello, world!"
\end{lstlisting}

This is more efficient than repeated \texttt{string-append} for building large strings.

\subsection{Input String Ports}
\index{open-input-string}

\begin{lstlisting}[style=repl]
> (define port (open-input-string "(+ 1 2) (+ 3 4)"))
> (read port)
(+ 1 2)
> (read port)
(+ 3 4)
> (read port)
#<eof>
\end{lstlisting}

Input string ports are useful for parsing structured data from strings.

\subsection{Capturing Output as a String}

A handy pattern using an output string port:

\begin{lstlisting}[style=repl]
> (define (value->string val)
    (let ((port (open-output-string)))
      (write val port)
      (get-output-string port)))
> (value->string '(1 2 3))
"(1 2 3)"
> (value->string "hello")
"\"hello\""
\end{lstlisting}

\section{The EOF Object}
\index{eof object}

When a read operation reaches the end of input, it returns an \emph{eof object}---a special value distinct from all other Scheme values:

\begin{lstlisting}[style=repl]
> (define port (open-input-string ""))
> (read-line port)
#<eof>
> (eof-object? (read-line port))
#t
\end{lstlisting}

Always test for EOF using \texttt{eof-object?}. The standard loop pattern is:

\begin{lstlisting}
(let loop ((datum (read port)))
  (unless (eof-object? datum)
    ;; process datum
    (loop (read port))))
\end{lstlisting}

\section{\texttt{read-char} and \texttt{write-char}}
\index{read-char}\index{write-char}

For character-level I/O:

\begin{lstlisting}[style=repl]
> (define port (open-input-string "hello"))
> (read-char port)
#\h
> (read-char port)
#\e
> (peek-char port)
#\l
> (read-char port)
#\l
\end{lstlisting}

\texttt{peek-char}\index{peek-char} returns the next character without consuming it---useful for lookahead in parsers.

\texttt{write-char} writes a single character to a port:

\begin{lstlisting}[style=repl]
> (write-char #\A)
A
> (write-char #\newline)

\end{lstlisting}

Between whole lines and single characters sit \texttt{read-string}\index{read-string} and \texttt{write-string}\index{write-string}. \texttt{(read-string k port)} reads up to \emph{k} characters and returns them as a string:

\begin{lstlisting}[style=repl]
> (define port (open-input-string "hello world"))
> (read-string 5 port)
"hello"
> (read-string 100 port)
" world"
\end{lstlisting}

\section{Binary Ports}
\index{binary ports}\index{bytevectors}

Every port seen so far is \emph{textual}: it reads and writes characters. R7RS also defines \emph{binary} ports, which carry raw bytes. Open them with \texttt{open-binary-input-file}\index{open-binary-input-file} and \texttt{open-binary-output-file}\index{open-binary-output-file}, and move data with \texttt{read-u8}\index{read-u8}, \texttt{write-u8}\index{write-u8}, \texttt{read-bytevector}\index{read-bytevector}, and \texttt{write-bytevector}\index{write-bytevector}:

\begin{lstlisting}
;; Read the first four bytes of a file (e.g. to check
;; a format signature)
(define (file-magic filename)
  (call-with-port (open-binary-input-file filename)
    (lambda (port)
      (read-bytevector 4 port))))
\end{lstlisting}

Real PNG files start with the signature bytes 137, 80, 78, 71. We can fabricate a minimal test file with \texttt{write-bytevector} and check it:

\begin{lstlisting}[style=repl]
> (call-with-port (open-binary-output-file "photo.png")
    (lambda (port)
      (write-bytevector (bytevector 137 80 78 71 13 10 26 10)
                        port)))
> (file-magic "photo.png")
#u8(137 80 78 71)
\end{lstlisting}

Binary reads return bytevectors (Chapter~\ref{ch:data-types}) or, at end of file, the same EOF object as textual reads. A port is either textual or binary, never both---calling \texttt{read-line} on a binary port is an error. Use binary ports for images, compressed data, and network protocols; use textual ports for anything meant to be read by humans.

\section{Putting It Together: A File Processing Example}

Here is a program that reads a text file, numbers each line, and writes the result to a new file:

\begin{lstlisting}
(define (number-lines input-file output-file)
  (call-with-port (open-input-file input-file)
    (lambda (in)
      (call-with-port (open-output-file output-file)
        (lambda (out)
          (let loop ((line (read-line in))
                     (n 1))
            (unless (eof-object? line)
              (display n out)
              (display ": " out)
              (display line out)
              (newline out)
              (loop (read-line in) (+ n 1)))))))))
\end{lstlisting}

Usage, with \texttt{write-lines} and \texttt{read-all-lines} from earlier in the chapter:

\begin{lstlisting}[style=repl]
> (write-lines "input.txt" '("alpha" "beta" "gamma"))
> (number-lines "input.txt" "output.txt")
> (for-each (lambda (l) (display l) (newline))
            (read-all-lines "output.txt"))
1: alpha
2: beta
3: gamma
\end{lstlisting}

\section*{Summary}

\begin{itemize}
    \item \texttt{display} produces human-readable output, while \texttt{write} produces machine-readable output that \texttt{read} can parse back.
    \item \texttt{read-line} reads a line of text and \texttt{read} reads a single Scheme datum.
    \item All I/O flows through ports; \texttt{call-with-port} and the \texttt{with-*-to-file} forms open and close them safely.
    \item String ports perform in-memory I/O, including capturing output as a string.
    \item Binary ports carry raw bytes as bytevectors; textual ports carry characters. A port is one or the other.
    \item Input procedures return the EOF object when input is exhausted.
\end{itemize}

\section*{Exercises}

\begin{exercise}[A Simple \code{cat}]
Write a function \code{cat} that takes a filename and prints its contents to the console, line by line. Use \code{call-with-port} to ensure the file is properly closed:

\begin{lstlisting}
(define (cat filename)
  ;; your code here
  )
\end{lstlisting}

Test it by creating a small text file and running \code{(cat "test.txt")} in the REPL. This is the Scheme equivalent of the Unix \texttt{cat} command (or Python's \texttt{print(open(f).read())}).
\end{exercise}

\begin{exercise}[Word Count]
Write a function \code{wc} that takes a filename and returns three values (using \code{values}): the number of lines, words, and characters in the file. Use \code{read-line} in a loop to process the file line by line:

\begin{lstlisting}[style=repl]
> (let-values (((lines words chars) (wc "example.txt")))
    (display lines) (display " lines, ")
    (display words) (display " words, ")
    (display chars) (display " characters")
    (newline))
5 lines, 23 words, 142 characters
\end{lstlisting}

Hint: for counting words in a line, use Kaappi's built-in \code{string-split} (a Kaappi extension to SRFI-13---always available, no import needed), or count transitions between whitespace and non-whitespace characters.
\end{exercise}

\begin{exercise}[File to List]
Write a function \code{file->lines} that reads all lines from a file and returns them as a list of strings. The chapter showed a \code{read-all-lines} helper---try writing it from memory:

\begin{lstlisting}[style=repl]
> (file->lines "example.txt")
("first line" "second line" "third line")
\end{lstlisting}

Then write the inverse: \code{lines->file} that takes a filename and a list of strings and writes each string as a line. Verify round-tripping: \code{(file->lines f)} after \code{(lines->file f data)} should return \code{data}.
\end{exercise}

\begin{exercise}[String Port Serialization]
Use string ports to implement \code{serialize} and \code{deserialize} functions that convert Scheme values to strings and back:

\begin{lstlisting}[style=repl]
> (define s (serialize '(1 "hello" #t (nested list))))
> s
"(1 \"hello\" #t (nested list))"
> (deserialize s)
(1 "hello" #t (nested list))
\end{lstlisting}

\code{serialize} should use \code{open-output-string} and \code{write}. \code{deserialize} should use \code{open-input-string} and \code{read}. Test with various types: numbers, strings, booleans, nested lists, and symbols. Why is \code{write} (not \code{display}) essential for round-tripping?
\end{exercise}

This completes Part~II of the book. In Part~III, we move to advanced topics, starting with Scheme's powerful macro system (Chapter~\ref{ch:macros}).

\chapterend
