\chapter{Macros}
\index{macros}\index{syntax-rules}
\label{ch:macros}

Macros are what make Lisp languages uniquely powerful. They let you extend the language itself---defining new syntactic forms that look and behave exactly like built-in ones. If you find yourself writing the same boilerplate pattern repeatedly, a macro can eliminate it.

Scheme's macro system is \emph{hygienic}\index{hygienic macros}: variables introduced by a macro never accidentally capture variables from the surrounding code. This eliminates an entire class of bugs that plague macro systems in C and Common Lisp.

This chapter covers \texttt{define-syntax} and \texttt{syntax-rules}---the pattern-based macro system specified by R7RS.

\section{What Is a Macro?}

A macro is a compile-time transformation rule\index{macro expansion}. When Kaappi encounters a macro call, it does not evaluate the arguments---instead, it rewrites the entire expression according to the macro's rules, and then evaluates the result.

One boundary is worth stating up front: macros transform \emph{expressions}, the tree of parenthesized forms the reader has already parsed. They cannot change how characters are read in the first place. Some Lisps let you extend the reader itself with new lexical syntax; Kaappi (like standard R7RS) does not---the token syntax of Chapter~\ref{ch:first-steps} is fixed.\index{reader}

Consider \texttt{when}:

\begin{lstlisting}[style=repl]
> (when (> 5 3)
    (display "yes")
    (newline))
yes
\end{lstlisting}

\texttt{when} is not a built-in---it is a macro that expands to an \texttt{if}:

\begin{lstlisting}
(if (> 5 3)
    (begin
      (display "yes")
      (newline)))
\end{lstlisting}

The macro takes the \emph{source code} and transforms it into different source code before evaluation:

\begin{center}
\begin{tikzpicture}[
    box/.style={rectangle, draw, rounded corners, font=\ttfamily\scriptsize,
                text width=5.8cm, align=left, inner sep=4pt},
    arr/.style={->, >=stealth, thick},
    lbl/.style={font=\sffamily\scriptsize, fill=white, inner sep=1pt}
]
\node[box] (src) {(when (> 5 3)\\~~(display "yes")\\~~(newline))};
\node[box, below=1.2cm of src] (exp) {(if (> 5 3)\\~~(begin\\~~~~(display "yes")\\~~~~(newline)))};
\draw[arr] (src) -- (exp) node[lbl, midway] {macro expansion};
\end{tikzpicture}
\end{center}

\begin{note}[Coming from Python]
Python has no equivalent feature. Decorators modify functions but cannot create new syntax. Macros operate at the level of code itself---they generate code from code.
\end{note}

\section{Your First Macro}

Define a macro with \texttt{define-syntax}\index{define-syntax} and \texttt{syntax-rules}:

\begin{lstlisting}[style=repl]
> (define-syntax my-when
    (syntax-rules ()
      ((my-when test body ...)
       (if test (begin body ...)))))
> (my-when (> 5 3)
    (display "hello")
    (newline))
hello
\end{lstlisting}

Let's break this down:

\begin{itemize}
    \item \texttt{define-syntax} names the macro (\texttt{my-when}).
    \item \texttt{syntax-rules ()} introduces the transformation rules. The \texttt{()} is a list of literal keywords (empty here).
    \item \texttt{(my-when test body ...)} is the \emph{pattern}---it matches the macro call.
    \item \texttt{(if test (begin body ...))} is the \emph{template}---what the macro expands to.
    \item \texttt{...} (ellipsis) means ``zero or more of the preceding element.''
\end{itemize}

\section{Seeing the Expansion}
\index{macro expansion!inspecting}\index{expand REPL command@\texttt{,expand} REPL command}

When a macro misbehaves, look at what it expands to. The Kaappi REPL's \texttt{,expand} command shows the expansion of any expression without evaluating it:

\begin{lstlisting}[style=repl]
> ,expand (when #t (display "yes"))
(if #t (begin (display "yes")))
> ,expand (my-when (> 5 3) (display "hello") (newline))
(if (> 5 3) (begin (display "hello") (newline)))
\end{lstlisting}

Make \texttt{,expand} a habit while working through this chapter: after each macro you write, expand a sample call and check that the generated code is what you intended. It turns macro debugging from guesswork into reading.

\section{Pattern Matching}

\texttt{syntax-rules} patterns work by structural matching:

\begin{itemize}
    \item A bare identifier (like \texttt{test} or \texttt{body}) matches any single expression and binds it as a \emph{pattern variable}.
    \item An identifier followed by \texttt{...} matches zero or more expressions.
    \item Literal parentheses match literal parentheses in the input.
    \item Identifiers listed in the literals list are matched by name, not as pattern variables.
    \item The underscore \texttt{\_}\index{underscore pattern} is a wildcard: it matches any single form without binding it. Because the macro name in a call always matches trivially, real-world macros conventionally write the name position as \texttt{\_}: \texttt{((\_ test body ...) ...)}. This book spells the name out for readability, but expect \texttt{\_} in code you read elsewhere.
\end{itemize}

\subsection{Multiple Clauses}

A macro can have multiple patterns. The first matching pattern wins:

\begin{lstlisting}[style=repl]
> (define-syntax my-and
    (syntax-rules ()
      ((my-and) #t)
      ((my-and x) x)
      ((my-and x rest ...)
       (if x (my-and rest ...) #f))))
> (my-and)
#t
> (my-and 1 2 3)
3
> (my-and 1 #f 3)
#f
\end{lstlisting}

This is a recursive macro: the third clause expands to another call to \texttt{my-and} with fewer arguments, until it reaches a base case.

If no clause matches a call, expansion fails with a compile-time syntax error at the call site. The \texttt{dotimes} macro later in this chapter shows how to catch a misuse deliberately with \texttt{syntax-error}.

\subsection{Literal Keywords}

The first argument to \texttt{syntax-rules} is a list of identifiers to match literally rather than as pattern variables:

\begin{lstlisting}[style=repl]
> (define-syntax my-if
    (syntax-rules (then else)
      ((my-if test then consequent else alternate)
       (cond (test consequent)
             (else alternate)))))
> (my-if (> 3 2) then "yes" else "no")
"yes"
\end{lstlisting}

Here, \texttt{then} and \texttt{else} are literal keywords---they must appear in the call exactly as written, not as variable names.

\section{The Ellipsis (\texttt{...})}
\index{ellipsis}

The ellipsis is the key to writing macros that accept variable numbers of subforms.

In a pattern, \texttt{expr ...} matches zero or more expressions. In a template, \texttt{expr ...} replicates the template for each matched element:

\begin{lstlisting}[style=repl]
> (define-syntax my-list
    (syntax-rules ()
      ((my-list) '())
      ((my-list elem rest ...)
       (cons elem (my-list rest ...)))))
> (my-list 1 2 3)
(1 2 3)
\end{lstlisting}

A more useful example---a macro that returns the value of the first expression while evaluating the rest for side effects:

\begin{lstlisting}[style=repl]
> (define-syntax begin0
    (syntax-rules ()
      ((begin0 first rest ...)
       (let ((result first))
         rest ...
         result))))
> (begin0 (+ 1 2)
    (display "side effect")
    (newline))
side effect
3
\end{lstlisting}

\subsection{Nested Ellipsis}

Ellipsis can be nested for macros that process lists of lists:

\begin{lstlisting}[style=repl]
> (define-syntax my-let
    (syntax-rules ()
      ((my-let ((var init) ...) body ...)
       ((lambda (var ...) body ...) init ...))))
> (my-let ((x 5) (y 10))
    (+ x y))
15
\end{lstlisting}

This \texttt{my-let} macro transforms into a lambda application---exactly how \texttt{let} is defined in the language.

\section{Hygiene}
\index{hygienic macros}

Scheme macros are hygienic: temporary variables introduced by a macro cannot clash with variables in the calling code. Consider:

\begin{lstlisting}[style=repl]
> (define-syntax swap!
    (syntax-rules ()
      ((swap! a b)
       (let ((tmp a))
         (set! a b)
         (set! b tmp)))))
> (let ((tmp 100) (x 1) (y 2))
    (swap! x y)
    (list tmp x y))
(100 2 1)
\end{lstlisting}

The \texttt{tmp} inside the macro and the \texttt{tmp} in the calling code are different bindings---the macro's \texttt{tmp} is automatically renamed to avoid capture\index{variable capture}. In an unhygienic system (like C's \texttt{\#define}), this would be a bug.

There is no magic here, and \texttt{,expand} lets you watch the mechanism work. During expansion, every identifier that comes from the \emph{template} (rather than from the macro call) is renamed to a fresh, globally unique name:

\begin{lstlisting}[style=repl]
> ,expand (swap! x y)
(__hyg_1_let ((__hyg_2_tmp x)) (set! x y) (set! y __hyg_2_tmp))
\end{lstlisting}

Read the expansion alongside the template. The template's \texttt{tmp} became \texttt{\_\_hyg\_2\_tmp}, so it can never collide with the \texttt{tmp} at the call site. The \texttt{x} and \texttt{y} came from the macro \emph{call}, not the template, so they keep their names and refer to the caller's bindings. Even \texttt{let} was renamed---the expander renames every template identifier uniformly, and the compiler recognizes renamed references to standard forms---which means the macro still works if the caller has shadowed \texttt{let} with a local variable. (The numbers in the fresh names are a session-wide counter, so yours will differ.)

Hygiene, then, is just consistent renaming: template identifiers are renamed away from the caller's namespace, call-site identifiers pass through untouched. \index{gensym}

\begin{note}[Why hygiene matters]
In C, the macro \texttt{\#define SWAP(a,b) \{int tmp=a; a=b; b=tmp;\}} breaks if you write \texttt{SWAP(tmp, x)} because the user's \texttt{tmp} and the macro's \texttt{tmp} collide. Scheme's system makes this impossible.
\end{note}

\section{Practical Macros}

\subsection{A \texttt{unless} Macro}

Both \texttt{when} and \texttt{unless} are built-in, but they are defined as macros internally. Here is how \texttt{unless} could be implemented:

\begin{lstlisting}
(define-syntax unless
  (syntax-rules ()
    ((unless test body ...)
     (if (not test) (begin body ...)))))
\end{lstlisting}

\subsection{A \texttt{while} Loop}
\index{while}

Scheme does not have a built-in \texttt{while}, but you can add one:

\begin{lstlisting}[style=repl]
> (define-syntax while
    (syntax-rules ()
      ((while test body ...)
       (let loop ()
         (when test
           body ...
           (loop))))))
> (let ((i 0))
    (while (< i 5)
      (display i) (display " ")
      (set! i (+ i 1))))
0 1 2 3 4
\end{lstlisting}

\subsection{A \texttt{dotimes} Loop}
\index{dotimes}

\begin{lstlisting}[style=repl]
> (define-syntax dotimes
    (syntax-rules ()
      ((dotimes (var count) body ...)
       (let loop ((var 0))
         (when (< var count)
           body ...
           (loop (+ var 1)))))))
> (dotimes (i 5)
    (display (* i i))
    (display " "))
0 1 4 9 16
\end{lstlisting}

A call that does not match the pattern---say \texttt{(dotimes 5 ...)}, forgetting the parentheses around \texttt{(i 5)}---fails with a syntax error at expansion time. You can make such misuse explicit by adding a catch-all clause that expands to \texttt{syntax-error}\index{syntax-error}, an R7RS form that signals an error during expansion:

\begin{lstlisting}
(define-syntax dotimes
  (syntax-rules ()
    ((dotimes (var count) body ...)
     (let loop ((var 0))
       (when (< var count) body ... (loop (+ var 1)))))
    ((dotimes other ...)
     (syntax-error "dotimes: expected (variable count)"))))
\end{lstlisting}

The catch-all clause matches anything the first clause rejected, and the \texttt{syntax-error} in its template stops compilation. The message string also documents, right in the macro definition, what shape of call was expected.

\subsection{An Assertion Macro}

\begin{lstlisting}[style=repl]
> (define-syntax assert
    (syntax-rules ()
      ((assert expr)
       (unless expr
         (error "Assertion failed" 'expr)))))
> (assert (= (+ 1 1) 2))
> (assert (= (+ 1 1) 3))
error[KP3000]: Assertion failed (= (+ 1 1) 3)
\end{lstlisting}

The failing assertion raises an error object whose message is \texttt{"Assertion failed"} and whose irritant is the \emph{source code} of the assertion---\texttt{'expr} in the template quotes it, which is why the report shows the exact expression that failed. A handler can read both parts back; Chapter~\ref{ch:errors} shows how.

\subsection{A Timing Macro}

This macro reports how long an expression takes to evaluate. \texttt{current-second}, from the standard \texttt{(scheme time)} library (Chapter~\ref{ch:stdlib}), returns the current time in seconds:

\begin{lstlisting}
(define-syntax time-it
  (syntax-rules ()
    ((time-it expr)
     (let ((start (current-second)))
       (let ((result expr))
         (let ((elapsed (- (current-second) start)))
           (display "Elapsed: ")
           (display elapsed)
           (display "s")
           (newline)
           result))))))
\end{lstlisting}

\section{Local Macros with \texttt{let-syntax}}
\index{let-syntax}\index{letrec-syntax}

You can define macros with limited scope using \texttt{let-syntax}:

\begin{lstlisting}[style=repl]
> (let-syntax ((double (syntax-rules ()
                         ((double x) (+ x x)))))
    (double 21))
42
\end{lstlisting}

The macro \texttt{double} exists only within the body. For mutually recursive macros, use \texttt{letrec-syntax}.

\begin{warning}[Macro arguments may be evaluated more than once]
A \texttt{syntax-rules} macro substitutes its arguments as \emph{code}, not as computed values. If a pattern variable appears several times in the template, its argument expression runs once for each occurrence. A \texttt{square} macro whose body is \texttt{(* x x)} expands \texttt{(square (read))} into \texttt{(* (read) (read))}, reading its input twice. When an argument may have side effects or is expensive, bind it once with \texttt{let} inside the macro before reusing it.
\end{warning}

\section{When to Use Macros}

Macros are powerful but should not be overused. Use a macro when:

\begin{itemize}
    \item You need to control \emph{when} or \emph{whether} arguments are evaluated (like \texttt{when}, \texttt{unless}, \texttt{and}).
    \item You need to introduce new bindings (like \texttt{let}, \texttt{dotimes}).
    \item You want to generate code at compile time to avoid runtime overhead.
    \item You are building a domain-specific language (DSL) with custom syntax.
\end{itemize}

Do \emph{not} use a macro when a function would suffice. Functions are simpler, composable (you can pass them to \texttt{map}), and easier to debug. If your ``macro'' does not need to delay evaluation or introduce bindings, make it a function.

\section*{Summary}

\begin{itemize}
    \item Macros transform code at expansion time, before evaluation, letting you add new syntactic forms.
    \item \texttt{syntax-rules} matches input against patterns and rewrites it using output templates.
    \item The ellipsis (\texttt{...}) matches and expands sequences of forms and can be nested.
    \item Scheme macros are hygienic, so introduced identifiers never accidentally capture the user's variables.
    \item The REPL's \texttt{,expand} command shows what a call expands to---the first tool to reach for when a macro misbehaves.
    \item \texttt{let-syntax} defines local macros; reach for macros only when a function cannot do the job.
\end{itemize}

\section*{Exercises}

\begin{exercise}[Swap!]
Write a \code{swap!} macro that swaps the values of two variables using \code{set!}. Verify that hygiene prevents the macro's temporary variable from clashing with user code.

\begin{lstlisting}[style=repl]
> (let ((a 1) (b 2))
    (swap! a b)
    (list a b))
(2 1)
> (let ((tmp 99) (x 10) (y 20))
    (swap! x y)
    (list tmp x y))
(99 20 10)
\end{lstlisting}
\end{exercise}

\begin{exercise}[Assert-Equal]
Write an \code{assert-equal} macro that takes two expressions, evaluates both, and raises an error showing the source code of each expression and their values if they are not \code{equal?}. (Hint: use \code{'expr} in the template to capture source code.)

\begin{lstlisting}[style=repl]
> (assert-equal (+ 1 2) 3)
;; no output --- assertion passed
> (assert-equal (* 2 3) 7)
error[KP3000]: assertion failed: (* 2 3) evaluated to 6, expected 7
\end{lstlisting}
\end{exercise}

\begin{exercise}[Do-Times with Named Counter]
Write a \code{do-n-times} macro similar to the \code{dotimes} macro from this chapter, but the user supplies the counter variable name, the count, \emph{and} an optional step size (defaulting to 1). You will need two \code{syntax-rules} clauses.

\begin{lstlisting}[style=repl]
> (do-n-times (i 5)
    (display i) (display " "))
0 1 2 3 4
> (do-n-times (i 10 3)
    (display i) (display " "))
0 3 6 9
\end{lstlisting}
\end{exercise}

\begin{exercise}[With-Logging]
Write a \code{with-logging} macro that prints an entry message before evaluating the body and an exit message after, then returns the body's result. The macro should accept a label string as its first argument.

\begin{lstlisting}[style=repl]
> (with-logging "compute"
    (+ 1 2))
[enter: compute]
[exit: compute]
3
\end{lstlisting}

As a bonus, make it print the elapsed time in the exit message using \code{current-second}.
\end{exercise}

In the next chapter (Chapter~\ref{ch:libraries}), we cover the library system---how to organize code into modules with clean interfaces.

\chapterend
