\chapter{The Library System}
\index{libraries}\index{modules}\index{define-library}
\label{ch:libraries}

As programs grow, you need to organize code into separate files with clean interfaces. Scheme's R7RS library system provides \texttt{define-library} for creating modules and \texttt{import} for using them. This chapter covers importing standard libraries, writing your own, controlling exports, and structuring multi-file projects.

\section{Importing Libraries}
\index{import}

Every Scheme program begins by importing the libraries it needs:

\begin{lstlisting}
(import (scheme base)
        (scheme write))

(display "Hello, world!")
(newline)
\end{lstlisting}

Strictly speaking, Kaappi does not force you to write these imports: every built-in procedure is globally available, in script files as well as in the REPL. Write them anyway. The import list documents which parts of the language a program uses, and other R7RS implementations \emph{require} it---code without imports is not portable. Imports also genuinely matter within Kaappi for the \emph{portable} SRFI libraries described below, which are unavailable until imported. You can import any number of libraries in a single \texttt{import} form. Notice that we import \texttt{(scheme write)} above: in R7RS, \texttt{display} and \texttt{write} are formally defined there, not in \texttt{(scheme base)}.

\subsection{Available Standard Libraries}

R7RS defines 16 standard libraries. The 15 you will actually use are listed below; the sixteenth, \texttt{(scheme r5rs)}, exists only for running legacy R5RS code:

\begin{center}
\begin{tabular}{lp{6cm}}
\textbf{Library} & \textbf{Contents} \\
\hline
\texttt{(scheme base)} & Core procedures: arithmetic, lists, strings, vectors, I/O, control \\
\texttt{(scheme write)} & \texttt{write}, \texttt{display}, \texttt{write-shared} \\
\texttt{(scheme read)} & \texttt{read} \\
\texttt{(scheme char)} & Unicode character procedures \\
\texttt{(scheme file)} & File port operations \\
\texttt{(scheme eval)} & \texttt{eval}, \texttt{environment} \\
\texttt{(scheme repl)} & \texttt{interaction-environment} \\
\texttt{(scheme load)} & \texttt{load} \\
\texttt{(scheme inexact)} & Transcendental functions (\texttt{sin}, \texttt{cos}, \texttt{sqrt}, etc.) \\
\texttt{(scheme complex)} & Complex number operations \\
\texttt{(scheme lazy)} & \texttt{delay}, \texttt{force}, promises \\
\texttt{(scheme time)} & \texttt{current-second}, jiffies \\
\texttt{(scheme case-lambda)} & \texttt{case-lambda} \\
\texttt{(scheme cxr)} & \texttt{cadr}, \texttt{caddr}, etc. (24 compositions) \\
\texttt{(scheme process-context)} & \texttt{exit}, \texttt{command-line}, environment variables \\
\end{tabular}
\end{center}

Beyond these, Kaappi provides 72 SRFIs importable as \texttt{(srfi N)}:

\begin{lstlisting}[style=repl]
> (import (srfi 1))   ; Extended list library
> (filter odd? '(1 2 3 4 5))
(1 3 5)
> (import (srfi 69))  ; Hash tables
> (import (srfi 13))  ; String library
\end{lstlisting}

The REPL also accepts the \texttt{,import} command, which takes the same forms: \texttt{,import (srfi 1)}.

\subsection{Built-in vs.\ Portable SRFIs}
\index{SRFI!built-in vs.\ portable}

The 72 SRFIs come in two kinds, and the difference shows up in practice:

\begin{itemize}
    \item \textbf{Eight built-in SRFIs} (1, 9, 13, 18, 39, 69, 133, 170) are implemented inside the runtime and always available---\texttt{filter} from SRFI-1 works even if you forget the import.
    \item \textbf{64 portable SRFIs} are ordinary \texttt{.sld} libraries loaded on demand. These \emph{must} be imported: \texttt{(list-sort < lst)} fails with an undefined-variable error until you write \texttt{(import (srfi 132))}.
\end{itemize}

When in doubt, write the import---it is always correct, and it keeps the program portable. Appendix~\ref{app:srfi} catalogs all 72 and marks which are built in.

\section{Import Modifiers}
\index{import modifiers}

Sometimes you want to import only specific names, avoid conflicts, or add a prefix. Import modifiers give you fine-grained control.

\begin{note}[Coming from Python]
Scheme's \texttt{import} brings names in \emph{unqualified} by default---closer to Python's \texttt{from module import *} than to \texttt{import module}. There is no \texttt{module.name} access; every imported binding becomes a plain top-level name. The modifiers below recover the control Python gives you: \texttt{only} is like \texttt{from m import a, b}; \texttt{rename} is like \texttt{from m import a as b}; and \texttt{prefix} gives you qualified-looking names such as \texttt{ht:ref}.
\end{note}

\subsection{\texttt{only}: Import Specific Names}
\index{only}

\begin{lstlisting}
(import (only (scheme base) map filter define let if))
\end{lstlisting}

Only the listed names become available. Everything else from the library is excluded.

\subsection{\texttt{except}: Exclude Specific Names}
\index{except}

\begin{lstlisting}
(import (except (scheme base) error))
\end{lstlisting}

Everything from \texttt{(scheme base)} is imported except \texttt{error}.

\subsection{\texttt{prefix}: Add a Namespace Prefix}
\index{prefix}

\begin{lstlisting}[style=repl]
> (import (prefix (scheme char) char:))
> (char:char-alphabetic? #\A)
#t
> (char:char-upcase #\a)
#\A
\end{lstlisting}

All imported names get the prefix \texttt{char:}. This avoids name collisions between libraries.

\subsection{\texttt{rename}: Rename Specific Imports}
\index{rename}

\begin{lstlisting}
(import (rename (scheme base) (define def) (lambda fn)))
(def square (fn (x) (* x x)))
\end{lstlisting}

Individual names can be renamed on import. This is useful when two libraries export the same name.

\subsection{Combining Modifiers}

Modifiers can be nested:

\begin{lstlisting}
(import (prefix (only (srfi 1) fold filter partition) list:))
;; Now available as: list:fold, list:filter, list:partition
\end{lstlisting}

\begin{warning}[Name conflicts are resolved silently]
If two imported libraries export the same name, Kaappi does not raise an error---the \emph{last} import wins, and the earlier binding is silently shadowed. That makes accidental clashes easy to miss. When two libraries you need both export the same name, make your intent explicit with \texttt{only}, \texttt{except}, \texttt{prefix}, or \texttt{rename}.
\end{warning}

\section{Writing Your Own Library}
\index{define-library}\index{.sld files}

A library is defined in a \texttt{.sld} file (Scheme Library Definition). The file name corresponds to the library name: the library \texttt{(mylib math)} lives in \texttt{mylib/math.sld}.

\subsection{Basic Structure}

\begin{lstlisting}
(define-library (mylib math)
  (export square cube factorial)
  (import (scheme base))
  (begin
    (define (square x) (* x x))

    (define (cube x) (* x x x))

    (define (factorial n)
      (let loop ((i n) (acc 1))
        (if (= i 0) acc
            (loop (- i 1) (* i acc)))))))
\end{lstlisting}

The key declarations:

\begin{itemize}
    \item \texttt{export}\index{export} --- lists the names that users of this library can access.
    \item \texttt{import} --- lists libraries that this library depends on.
    \item \texttt{begin} --- contains the Scheme code that defines the exported bindings.
\end{itemize}

\subsection{Using Your Library}

Create a program that imports it:

\begin{lstlisting}
(import (mylib math))

(display (square 5))
(newline)
(display (factorial 10))
(newline)
\end{lstlisting}

Run from the directory containing the \texttt{mylib/} folder:

\begin{Verbatim}[fontsize=\small]
kaappi program.scm
\end{Verbatim}

\begin{codeoutput}
25
3628800
\end{codeoutput}

\section{Library Search Paths}
\index{library search path}\index{--lib-path}

Kaappi searches for \texttt{.sld} files in this order:

\begin{enumerate}
    \item The current working directory (\texttt{./})
    \item The \texttt{./lib/} subdirectory
    \item Directories specified with \texttt{--lib-path} (repeatable, up to 16)
    \item The directory containing the script being run
    \item The user library directory (\texttt{\textasciitilde/.kaappi/lib/})
\end{enumerate}

The library name \texttt{(mylib utils)} maps to the file path \texttt{mylib/utils.sld}. Each component of the library name becomes a directory, except the last which becomes the filename.

To add custom search paths:

\begin{Verbatim}[fontsize=\small]
kaappi --lib-path ./vendor/libs --lib-path ./mylibs program.scm
\end{Verbatim}

\section{Export Specifications}

\subsection{Simple Exports}

\begin{lstlisting}
(export square cube factorial)
\end{lstlisting}

Each listed name becomes available to importers.

\subsection{Renamed Exports}

You can export a binding under a different name:

\begin{lstlisting}
(define-library (mylib strings)
  (export (rename internal-join join)
          (rename internal-split split))
  (import (scheme base))
  (begin
    (define (internal-join lst sep) ...)
    (define (internal-split s sep) ...)))
\end{lstlisting}

Users import \texttt{join} and \texttt{split}, but internally the functions are named \texttt{internal-join} and \texttt{internal-split}.

\subsection{Re-Exporting Imported Bindings}
\index{re-export}

A library can export names it imported from another library. This supports a \emph{facade}: one library presenting a single interface assembled from several internal ones:

\begin{lstlisting}
(define-library (mylib all)
  (export parse tokenize join split)
  (import (scheme base)
          (mylib parser)      ; provides parse, tokenize
          (mylib strings)))   ; provides join, split
\end{lstlisting}

Users import \texttt{(mylib all)} and get the combined interface, while you remain free to reorganize the internal libraries behind it.

\section{The \texttt{include} Declaration}
\index{include}

For libraries with large implementations, you can keep the code in separate files:

\begin{lstlisting}
(define-library (mylib parser)
  (export parse tokenize)
  (import (scheme base) (scheme char))
  (include "parser-impl.scm"))
\end{lstlisting}

The file \texttt{parser-impl.scm} is read as if its contents were inside a \texttt{begin} block. This keeps \texttt{.sld} files short and focused on the interface.

\section{\texttt{cond-expand}: Conditional Code}
\index{cond-expand}

\texttt{cond-expand} lets you include different code based on which features are available:

\begin{lstlisting}
(define-library (mylib platform)
  (export get-os)
  (import (scheme base))
  (cond-expand
    (kaappi
     (begin
       (define (get-os) "kaappi")))
    (else
     (begin
       (define (get-os) "unknown")))))
\end{lstlisting}

Kaappi defines the feature identifiers \texttt{kaappi}, \texttt{r7rs}, \texttt{ieee-float}, \texttt{posix}, and \texttt{exact-closed}. The built-in procedure \texttt{(features)}\index{features} returns the list the running interpreter reports:

\begin{lstlisting}[style=repl]
> (features)
(r7rs kaappi ieee-float posix)
\end{lstlisting}

\section{Project Structure}

A well-organized Kaappi project looks like this:

\begin{Verbatim}[fontsize=\small]
my-project/
  main.scm              # Entry point
  mylib/
    core.sld            # (mylib core)
    utils.sld           # (mylib utils)
    http/
      client.sld        # (mylib http client)
      server.sld        # (mylib http server)
\end{Verbatim}

\begin{lstlisting}
;; main.scm
(import (mylib core)
        (mylib utils)
        (mylib http client))

(start-app)
\end{lstlisting}

\subsection{Avoiding Circular Dependencies}

Libraries cannot have circular imports. If library A imports library B, then B cannot import A. If you encounter this, extract the shared definitions into a third library C that both A and B import.

\section{Packages with Thottam}
\index{thottam}

For sharing libraries with others, Kaappi uses the \texttt{thottam} package manager. A package is a library (or set of libraries) published to the package registry:

\begin{Verbatim}[fontsize=\small]
thottam install kaappi-json
\end{Verbatim}

After installation, you can import the library directly:

\begin{lstlisting}
(import (kaappi json))
\end{lstlisting}

Packages are installed to \texttt{\textasciitilde/.kaappi/lib/}, which is always on the search path. We will use several ecosystem packages in Part~IV. For reference material, see Appendix~\ref{app:cli} (the full \texttt{thottam} command set) and Appendix~\ref{app:ecosystem} (the catalog of available packages).

\section*{Summary}

\begin{itemize}
    \item \texttt{import} brings library bindings into scope, and the modifiers \texttt{only}, \texttt{except}, \texttt{prefix}, and \texttt{rename} control exactly what is imported.
    \item Kaappi's built-ins (including eight built-in SRFIs) are globally available, but the 64 portable SRFIs and your own libraries genuinely require \texttt{import}---and portable R7RS code always writes its imports.
    \item \texttt{define-library} declares a library with explicit \texttt{export} and \texttt{import} sets.
    \item Libraries are found through the search path, and \texttt{include} and \texttt{cond-expand} support file inclusion and conditional code.
    \item The \texttt{thottam} package manager publishes and installs third-party packages.
\end{itemize}

\section*{Exercises}

\begin{exercise}[Create a Utility Library]
Create a file \code{mylib/utils.sld} that defines a library \code{(mylib utils)} exporting three utility procedures: \code{clamp} (restricts a number to a min/max range), \code{repeat-string} (returns a string repeated \emph{n} times), and \code{displayln} (displays a value followed by a newline). Write a short program that imports and uses all three.

\begin{lstlisting}
;; mylib/utils.sld
(define-library (mylib utils)
  (export clamp repeat-string displayln)
  (import (scheme base) (scheme write))
  (begin
    ;; your definitions here
    ))
\end{lstlisting}

Test in the REPL:
\begin{lstlisting}[style=repl]
> (clamp 15 0 10)
10
> (repeat-string "ha" 3)
"hahaha"
\end{lstlisting}
\end{exercise}

\begin{exercise}[Import Modifiers]
Import SRFI-1 (the extended list library) three different ways in a single program using import modifiers:

\begin{enumerate}
    \item Use \code{only} to import just \code{fold} and \code{unfold}.
    \item Use \code{prefix} to import the entire library with a \code{list:} prefix.
    \item Use \code{rename} to import \code{partition} as \code{split-by}.
\end{enumerate}

Write a short program that demonstrates all three imports working. Which approach would you choose for a large project, and why?
\end{exercise}

\begin{exercise}[Conditional Code with cond-expand]
Write a library \code{(mylib greeting)} that uses \code{cond-expand} to export a \code{greet} procedure. When the \code{kaappi} feature is present, \code{greet} should mention it; on other Scheme implementations it should return a generic greeting. As a bonus, add an \code{(and kaappi posix)} branch that also includes the login name from \code{(get-environment-variable "USER")}.

\begin{lstlisting}[style=repl]
> (import (mylib greeting))
> (greet "Alice")
"Hello, Alice! (running on Kaappi, user: baiju)"
\end{lstlisting}
\end{exercise}

In the next chapter (Chapter~\ref{ch:records}), we define custom data types with records.

\chapterend
