\chapter{Error Handling}
\index{error handling}\index{exceptions}

Programs encounter errors: files that do not exist, invalid user input, network failures, division by zero. Scheme provides a structured exception system with \texttt{guard} for catching errors, \texttt{raise} for signaling them, and \texttt{error} for creating error objects. This chapter covers all three, plus strategies for writing robust code.

\section{The \texttt{error} Procedure}
\index{error}

The simplest way to signal an error is the \texttt{error} procedure:

\begin{lstlisting}[style=repl]
> (error "something went wrong")
; Error: something went wrong
> (error "index out of range" 5)
; Error: index out of range 5
> (error "expected a string" 42 'got-number)
; Error: expected a string 42 got-number
\end{lstlisting}

\texttt{error} creates an \emph{error object}\index{error object} and raises it as an exception. The first argument is a message string; additional arguments are \emph{irritants}---values that provide context about what went wrong.

\subsection{Using \texttt{error} in Your Functions}

\begin{lstlisting}[style=repl]
> (define (safe-divide a b)
    (if (zero? b)
        (error "division by zero" a b)
        (/ a b)))
> (safe-divide 10 2)
5
> (safe-divide 10 0)
; Error: division by zero 10 0
\end{lstlisting}

\begin{note}[Coming from Python]
\texttt{(error "message" value)} is like Python's \texttt{raise ValueError(f"message: \{value\}")}. The key difference is that Scheme's error objects are structured---the message and irritants are separate fields, not formatted into a string.
\end{note}

\section{Catching Errors with \texttt{guard}}
\index{guard}

\texttt{guard} is Scheme's primary exception-handling form. It catches raised exceptions and matches them against clauses, similar to \texttt{try}/\texttt{except} in Python:

\begin{lstlisting}[style=repl]
> (guard (e (#t (string-append "caught: "
                               (error-object-message e))))
    (error "file not found" "data.txt"))
"caught: file not found"
\end{lstlisting}

The general form:

\begin{lstlisting}
(guard (variable
        (test1 expr1 ...)
        (test2 expr2 ...)
        ...
        (else expr ...))
  body ...)
\end{lstlisting}

The \texttt{body} is evaluated. If an exception is raised, it is bound to \texttt{variable}, and the clauses are tested in order (like \texttt{cond}). The first matching clause's expressions are evaluated and returned. If no clause matches, the exception is re-raised.

\subsection{Multiple Clauses}

\begin{lstlisting}[style=repl]
> (guard (e
          ((string? e)
           (string-append "string error: " e))
          ((number? e)
           (* e 10))
          (else "unknown error"))
    (raise "oops"))
"string error: oops"
\end{lstlisting}

\subsection{Inspecting Error Objects}

Error objects created by \texttt{error} can be examined with three procedures:

\begin{lstlisting}[style=repl]
> (guard (e (#t (list
                  (error-object? e)
                  (error-object-message e)
                  (error-object-irritants e))))
    (error "bad value" 42 'expected-string))
(#t "bad value" (42 expected-string))
\end{lstlisting}

\begin{itemize}
    \item \texttt{error-object?}\index{error-object?} --- tests if the value is an error object
    \item \texttt{error-object-message}\index{error-object-message} --- extracts the message string
    \item \texttt{error-object-irritants}\index{error-object-irritants} --- extracts the list of irritant values
\end{itemize}

\subsection{Conditional Error Handling}

You can match on the error message or type to handle different errors differently:

\begin{lstlisting}
(define (safe-lookup key table)
  (guard (e
          ((and (error-object? e)
                (string=? (error-object-message e)
                          "key not found"))
           #f)
          (else (raise e)))  ; re-raise unknown errors
    (lookup key table)))
\end{lstlisting}

The \texttt{(else (raise e))} clause ensures that unexpected errors are not silently swallowed.

\begin{warning}[Always re-raise unknown errors]
Do not write \texttt{(guard (e (\#t \#f)) ...)} to blindly catch everything. This hides bugs. Only catch the specific errors you know how to handle, and re-raise the rest.
\end{warning}

\section{Raising Exceptions with \texttt{raise}}
\index{raise}

\texttt{raise} signals an exception. Any value can be raised---it does not have to be an error object:

\begin{lstlisting}[style=repl]
> (guard (e (#t e))
    (raise "a string error"))
"a string error"
> (guard (e (#t e))
    (raise 42))
42
> (guard (e (#t e))
    (raise '(custom error data)))
(custom error data)
\end{lstlisting}

However, the convention is to use \texttt{error} (which creates a proper error object) for most situations. Direct \texttt{raise} with arbitrary values is useful for custom condition types.

\subsection{\texttt{raise} vs.\ \texttt{error}}

\begin{center}
\begin{tabular}{lp{7cm}}
\textbf{Form} & \textbf{Use when} \\
\hline
\texttt{(error msg irritants ...)} & Signaling an error with a message---the common case \\
\texttt{(raise value)} & Signaling a non-error condition, or raising a custom condition object \\
\end{tabular}
\end{center}

\section{\texttt{with-exception-handler}: Low-Level Handler}
\index{with-exception-handler}

\texttt{with-exception-handler} installs an exception handler for the duration of a thunk:

\begin{lstlisting}[style=repl]
> (call/cc
    (lambda (exit)
      (with-exception-handler
        (lambda (e) (exit (list 'caught e)))
        (lambda () (raise "boom")))))
(caught "boom")
\end{lstlisting}

This is lower-level than \texttt{guard}. The handler receives the exception and must decide what to do: it can invoke an escape continuation, re-raise, or (for continuable exceptions) return a value.

In most code, \texttt{guard} is preferred because it is clearer and handles re-raising automatically. Use \texttt{with-exception-handler} only when you need the full power (e.g., logging all exceptions without catching them).

\section{Continuable Exceptions}
\index{raise-continuable}

\texttt{raise-continuable} raises an exception that the handler can ``fix'' by returning a value:

\begin{lstlisting}[style=repl]
> (with-exception-handler
    (lambda (e) (* e 10))    ; handler returns a replacement
    (lambda () (+ 1 (raise-continuable 5))))
51
\end{lstlisting}

The handler receives 5, returns 50, and execution continues as if \texttt{raise-continuable} had returned 50. This is an advanced feature---use it for situations where a handler can supply a default or corrected value.

\section{Common Patterns}

\subsection{Default Values on Error}

\begin{lstlisting}
(define (read-config-or-default filename)
  (guard (e (#t '()))  ; return empty list on any error
    (call-with-port (open-input-file filename)
      (lambda (port) (read port)))))
\end{lstlisting}

\subsection{Cleanup with \texttt{dynamic-wind}}
\index{dynamic-wind}

\texttt{dynamic-wind} ensures cleanup code runs even if an exception occurs:

\begin{lstlisting}
(define (with-temp-file proc)
  (let ((filename (make-temp-name)))
    (dynamic-wind
      (lambda () #f)                    ; before
      (lambda () (proc filename))       ; body
      (lambda () (delete-file filename))))) ; after (cleanup)
\end{lstlisting}

The ``after'' thunk runs whether the body returns normally or an exception is raised. This is similar to Python's \texttt{try}/\texttt{finally}.

\subsection{Wrapping External Calls}

\begin{lstlisting}
(define (safe-json-parse str)
  (guard (e
          ((error-object? e)
           (values #f (error-object-message e)))
          (else
           (values #f "unknown parse error")))
    (values (json-read (open-input-string str)) #f)))
\end{lstlisting}

This returns two values: the parsed result (or \texttt{\#f} on error) and an error message (or \texttt{\#f} on success).

\subsection{Asserting Preconditions}

\begin{lstlisting}
(define (vector-median v)
  (when (zero? (vector-length v))
    (error "cannot compute median of empty vector" v))
  (let ((sorted (vector-sort < v)))
    (vector-ref sorted (quotient (vector-length sorted) 2))))
\end{lstlisting}

Fail early with a clear message when preconditions are not met.

\section{Error Handling Strategy}

\begin{itemize}
    \item \textbf{Fail fast.} Use \texttt{error} to signal problems as soon as they are detected. Do not pass invalid data deeper into the program hoping something will work.
    \item \textbf{Catch at boundaries.} Handle errors at the boundary between subsystems---at the top of a request handler, at the entry point of a CLI tool, at the edge of a library call. Do not pepper every function with \texttt{guard}.
    \item \textbf{Be specific.} Catch only the errors you can handle. Let unexpected errors propagate---they usually indicate bugs that should not be silently swallowed.
    \item \textbf{Provide context.} Use irritants in \texttt{error} to include the values that caused the problem. ``index out of range'' is less useful than ``index out of range: 10, list length: 5.''
\end{itemize}

\sectionbreak

You now know how to signal errors with \texttt{error}, catch them with \texttt{guard}, raise custom conditions with \texttt{raise}, and write robust code that fails gracefully. In the next chapter, we explore continuations---Scheme's most powerful and unusual control flow mechanism.
