\chapter{The Standard Library}
\index{SRFI}\index{standard library}
\label{ch:stdlib}

Beyond the R7RS standard libraries (covered in Chapter~\ref{ch:libraries}), Kaappi provides 72 SRFIs (Scheme Requests for Implementation)---community-standard libraries that extend Scheme's capabilities. This chapter surveys the most useful SRFIs, showing practical examples of each. For the complete catalog of all 72 SRFIs Kaappi supports, see Appendix~\ref{app:srfi}.

\section{What Are SRFIs?}

SRFIs are standardized library specifications for Scheme. Each SRFI has a number, a title, and a precise specification. They are the Scheme equivalent of Python's PEPs or JavaScript's TC39 proposals---except that each SRFI defines a concrete library you can import and use immediately.

Import any SRFI by number:

\begin{lstlisting}
(import (srfi 1))   ; List library
(import (srfi 69))  ; Hash tables
(import (srfi 13))  ; String library
\end{lstlisting}

Kaappi's 72 SRFIs fall into two categories:

\begin{itemize}
    \item \textbf{8 built-in SRFIs} --- implemented in Zig for performance, always available.
    \item \textbf{64 portable SRFIs} --- pure Scheme implementations loaded on demand from \texttt{.sld} files.
\end{itemize}

In practice the split is invisible: both kinds are imported with the same \texttt{(import (srfi N))} form. The built-in ones are simply faster.

\section{SRFI-1: Extended List Library}
\index{SRFI-1}

SRFI-1 is the most commonly used extension. It provides dozens of list procedures beyond what \texttt{(scheme base)} offers:\index{iota}

\begin{lstlisting}[style=repl]
> (import (srfi 1))
> (iota 5)
(0 1 2 3 4)
> (iota 5 1)
(1 2 3 4 5)
> (iota 5 0 2)
(0 2 4 6 8)
\end{lstlisting}

\subsection{Searching and Filtering}
\index{find}\index{any}\index{every}\index{partition}

\begin{lstlisting}[style=repl]
> (find even? '(1 3 5 4 7))
4
> (any even? '(1 3 5 7))
#f
> (any even? '(1 3 4 7))
#t
> (every positive? '(1 2 3 4))
#t
> (partition even? '(1 2 3 4 5 6))
(2 4 6)
(1 3 5)
\end{lstlisting}

\subsection{Folding and Reducing}
\index{reduce}\index{unfold}

\begin{lstlisting}[style=repl]
> (fold + 0 '(1 2 3 4 5))
15
> (fold-right cons '() '(a b c))
(a b c)
> (reduce max 0 '(3 7 2 9 4))
9
> (unfold (lambda (x) (> x 5))
          (lambda (x) (* x x))
          (lambda (x) (+ x 1))
          1)
(1 4 9 16 25)
\end{lstlisting}

\subsection{Zipping and Unzipping}
\index{zip}

\begin{lstlisting}[style=repl]
> (zip '(a b c) '(1 2 3))
((a 1) (b 2) (c 3))
> (unzip2 '((a 1) (b 2) (c 3)))
(a b c)
(1 2 3)
\end{lstlisting}

\subsection{Other Highlights}
\index{take}\index{drop}\index{delete-duplicates}\index{concatenate}

\begin{lstlisting}[style=repl]
> (take '(a b c d e) 3)
(a b c)
> (drop '(a b c d e) 3)
(d e)
> (concatenate '((1 2) (3 4) (5 6)))
(1 2 3 4 5 6)
> (delete-duplicates '(1 2 3 2 1 4))
(1 2 3 4)
\end{lstlisting}

\section{SRFI-13: String Library}
\index{SRFI-13}

SRFI-13 extends Scheme's string operations significantly:

\begin{lstlisting}[style=repl]
> (import (srfi 13))
> (string-contains "hello world" "world")
6
> (string-prefix? "he" "hello")
#t
> (string-suffix? "ld" "world")
#t
> (string-trim "  hello  ")
"hello  "
> (string-trim-both "  hello  ")
"hello"
> (string-join '("one" "two" "three") ", ")
"one, two, three"
> (string-split "a,b,c,d" ",")
("a" "b" "c" "d")
\end{lstlisting}

\begin{note}[Coming from Python]
\texttt{string-join} is Python's
\texttt{", ".join([...])}.
\texttt{string-split} is Python's
\texttt{"a,b,c,d".split(",")}.
\end{note}

One portability caveat: \texttt{string-split} is a Kaappi extension to \texttt{(srfi 13)}. The SRFI itself defines \texttt{string-tokenize} instead, so code that uses \texttt{string-split} may need adjusting on other Schemes.

\section{SRFI-69: Hash Tables}
\index{SRFI-69}\index{hash tables}

Hash tables provide O(1) average key lookup---much faster than association lists for large collections:\index{make-hash-table}\index{hash-table-set"!}\index{hash-table-ref}\index{hash-table-ref/default}

\begin{lstlisting}[style=repl]
> (import (srfi 69))
> (define ht (make-hash-table))
> (hash-table-set! ht "name" "Alice")
> (hash-table-set! ht "age" 30)
> (hash-table-ref ht "name")
"Alice"
> (hash-table-ref ht "age")
30
> (hash-table-exists? ht "email")
#f
> (hash-table-ref/default ht "email" "unknown")
"unknown"
\end{lstlisting}

\begin{warning}[Keys always compare with \texttt{equal?}]
\index{hash tables!key equality}Kaappi accepts SRFI-69's optional equality and hash arguments to \texttt{make-hash-table} but ignores them: every table compares keys with \texttt{equal?}. Strings, lists, and vectors therefore work as keys by value, as the examples above show. Records do not---a record is \texttt{equal?} only to itself (Chapter~\ref{ch:records}), so a record key can be retrieved only with the very same object. To index by a record's contents, use a key built from plain data, such as a list of its field values.
\end{warning}

\subsection{Iteration}

\begin{lstlisting}[style=repl]
> (hash-table-set! ht "city" "London")
> (hash-table-keys ht)
("age" "city" "name")
> (hash-table-values ht)
(30 "London" "Alice")
> (hash-table-walk ht
    (lambda (key value)
      (display key) (display ": ")
      (display value) (newline)))
name: Alice
city: London
age: 30
\end{lstlisting}

Notice that the iteration order is \emph{unspecified}---it matches neither the insertion order nor each other across calls. Never rely on the order in which a hash table yields its entries; sort the keys first if you need a stable order.

\subsection{Building from Data}

To load a hash table from existing data, such as an association list, iterate over the entries:

\begin{lstlisting}[style=repl]
> (define color-alist '(("red" . "#ff0000")
                        ("green" . "#00ff00")
                        ("blue" . "#0000ff")))
> (define ht (make-hash-table))
> (for-each (lambda (entry)
              (hash-table-set! ht (car entry) (cdr entry)))
            color-alist)
> (hash-table-ref ht "green")
"#00ff00"
\end{lstlisting}

\section{SRFI-133: Vector Library}
\index{SRFI-133}

SRFI-133 adds higher-order operations on vectors:

\begin{lstlisting}[style=repl]
> (import (srfi 133))
> (vector-map + #(1 2 3) #(10 20 30))
#(11 22 33)
> (vector-count even? #(1 2 3 4 5 6))
3
> (vector-any odd? #(2 4 5 6))
#t
> (vector-every positive? #(1 2 3 4))
#t
> (vector-index even? #(1 3 4 7))
2
\end{lstlisting}

\section{SRFI-132: Sorting}
\index{SRFI-132}\index{sorting}\index{list-sort}\index{vector-sort}

\begin{lstlisting}[style=repl]
> (import (srfi 132))
> (list-sort < '(5 3 1 4 2))
(1 2 3 4 5)
> (list-sort string<? '("banana" "apple" "cherry"))
("apple" "banana" "cherry")
> (vector-sort < #(5 3 1 4 2))
#(1 2 3 4 5)
\end{lstlisting}

The comparator is a two-argument predicate. Use \texttt{<} for ascending numbers, \texttt{>} for descending, \texttt{string<?} for strings.

\section{SRFI-27: Random Numbers}
\index{SRFI-27}

\begin{lstlisting}[style=repl]
> (import (srfi 27))
> (random-integer 100)
42    (*@\emph{(your numbers will differ)}@*)
> (random-integer 100)
73
> (random-real)
0.6234981723
\end{lstlisting}

\texttt{random-integer}\index{random-integer} returns a random integer in \texttt{[0, n)}. \texttt{random-real}\index{random-real} returns a random real in \texttt{[0, 1)}.

\section{SRFI-115: Regular Expressions}
\index{SRFI-115}\index{regular expressions}

\begin{lstlisting}[style=repl]
> (import (srfi 115))
> (regexp-matches '(: "hello") "hello world")
#f
> (regexp-search '(: "world") "hello world")
#<<regexp-match> "hello world" #((6 11))>
> (regexp-replace '(+ num) "age: 25" "XX")
"age: XX"
> (regexp-extract '(+ alpha) "foo123bar456baz")
("foo" "bar" "baz")
\end{lstlisting}

SRFI-115 uses S-expression syntax for patterns instead of string-based regex. \texttt{(+ num)} matches one or more digits, \texttt{(+ alpha)} matches one or more letters. A match object records the searched string and the span of each match---here characters 6 to 11.\index{regexp-matches}\index{regexp-replace}\index{regexp-extract}

\section{SRFI-158: Generators}
\index{SRFI-158}\index{generators}

Generators are lazy sequences---they produce values one at a time on demand:\index{make-range-generator}\index{generator->list}\index{gfilter}

\begin{lstlisting}[style=repl]
> (import (srfi 158))
> (define g (make-range-generator 1 6))
> (g)
1
> (g)
2
> (g)
3
> (generator->list (make-range-generator 0 5))
(0 1 2 3 4)
> (generator->list
    (gfilter even? (make-range-generator 0 10)))
(0 2 4 6 8)
\end{lstlisting}

Generators are useful for processing large or infinite sequences without materializing them all in memory.

\section{SRFI-64: Unit Testing}
\index{SRFI-64}\index{testing}\index{unit testing}\index{test-equal}\index{test-assert}

Every project you keep needs tests, and Scheme's standard test framework is SRFI-64. You group related checks between \texttt{test-begin} and \texttt{test-end}, and state expectations with \texttt{test-equal} (compares with \texttt{equal?}) and \texttt{test-assert} (expects a true value). Each test takes an optional descriptive name:

\begin{lstlisting}[style=repl]
> (import (srfi 64))
> (test-begin "arith")
%%%% Starting test arith  (Writing full log to "arith.log")
> (test-equal "two plus two" 4 (+ 2 2))
> (test-assert "ten is even" (even? 10))
> (test-equal "off by one" 5 (+ 2 2))
FAIL off by one
> (test-end "arith")
# of expected passes      2
# of unexpected failures  1
\end{lstlisting}

Passing tests are silent. Failures are reported the moment they run, and \texttt{test-end} prints the tally for the group. The full record---every test's name, actual value, and expected value---is written to the log file named after the group (here \texttt{arith.log}), which is where you look when a failure message alone is not enough.

\texttt{test-error} asserts that an expression \emph{raises} an error, which is how you test the failure paths of Chapter~\ref{ch:errors}:

\begin{lstlisting}[style=repl]
> (test-begin "errors")
%%%% Starting test errors  (Writing full log to "errors.log")
> (test-error "car of non-pair" (car 5))
> (test-end "errors")
# of expected passes      1
\end{lstlisting}

In a real project, tests live in their own files---the ecosystem convention is a \texttt{tests/} directory next to \texttt{lib/}---and each file is an ordinary script you run with \texttt{kaappi tests/test-mylib.scm}. Import the library under test, exercise its exports, and let the summary tell you where you stand. To run a whole suite at once, hand the directory to the built-in runner: \texttt{kaappi test tests/}\index{kaappi test} executes each file in its own subprocess, aggregates the pass/fail tallies, and exits non-zero on any failure---exactly what a CI job wants. Kaappi can also report which exported procedures your tests never called: run the suite with \texttt{kaappi --coverage} (Appendix~\ref{app:cli}).

If SRFI-64 feels heavyweight for a quick script, SRFI-78\index{SRFI-78} offers a one-form alternative: \texttt{check} compares an expression against an expected value and reports immediately:

\begin{lstlisting}[style=repl]
> (import (srfi 78))
> (check (+ 2 2) => 4)
(1) (+ 2 2) => 4 ; correct
> (check-report)

Checks: 1 total, 1 passed, 0 failed.
\end{lstlisting}

Beyond the SRFIs, the ecosystem provides \texttt{kaappi-test} (a richer test framework with suites and CI-friendly output) and \texttt{kaappi-bdd} (behavior-driven \texttt{describe}/\texttt{it}/\texttt{expect} style); see Appendix~\ref{app:ecosystem}.

\section{Other Useful SRFIs}

\subsection{SRFI-17: Generalized \texttt{set!}}
\index{SRFI-17}\index{generalized set"!}

With SRFI-17, \texttt{set!} accepts an \emph{accessor call} as its target and updates the place that accessor reads---the closest Scheme comes to Python's \texttt{p[0] = x}:

\begin{lstlisting}[style=repl]
> (import (srfi 17))
> (define p (list 1 2 3))
> (set! (car p) 99)
> p
(99 2 3)
> (define v (vector 'a 'b 'c))
> (set! (vector-ref v 2) 'z)
> v
#(a b z)
\end{lstlisting}

Setters come pre-defined for the common accessors, including \texttt{car}, \texttt{cdr}, \texttt{vector-ref}, and \texttt{string-ref}---\texttt{(set! (vector-ref v 2) 'z)} means \texttt{(vector-set! v 2 'z)}. The mutation caveats of Chapter~\ref{ch:lists} still apply; this changes the spelling, not the semantics.

\subsection{SRFI-61: A More General \texttt{cond}}
\index{SRFI-61}\index{cond}

Chapter~\ref{ch:control-flow}'s \texttt{cond} can pass a truthy test value to a receiver with \texttt{=>}. SRFI-61 extends the clause to \texttt{(generator guard => receiver)}: the generator produces a value, the \emph{guard} decides whether the clause fires, and the receiver gets the value. That separates ``produced a value'' from ``produced the right kind of value'':

\begin{lstlisting}[style=repl]
> (import (srfi 61))
> (define (classify s)
    (cond ((string->number s) exact-integer? =>
           (lambda (n) (list 'integer n)))
          ((string->number s) number? =>
           (lambda (n) (list 'number n)))
          (else 'not-a-number)))
> (classify "42")
(integer 42)
> (classify "3.5")
(number 3.5)
> (classify "hello")
not-a-number
\end{lstlisting}

With plain \texttt{cond}, the first clause would fire for \emph{any} number; the \texttt{exact-integer?} guard lets \texttt{"3.5"} fall through to the second clause without parsing the string twice by hand.

\subsection{SRFI-26: Partial Application with \texttt{cut}}
\index{SRFI-26}\index{cut}

\begin{lstlisting}[style=repl]
> (import (srfi 26))
> (map (cut + <> 10) '(1 2 3))
(11 12 13)
> (map (cut string-append "prefix-" <>) '("a" "b" "c"))
("prefix-a" "prefix-b" "prefix-c")
\end{lstlisting}

\texttt{cut} creates a partial application. \texttt{<>} marks the argument slot.

\subsection{SRFI-28: Basic Format Strings}
\index{SRFI-28}\index{format}

\begin{lstlisting}[style=repl]
> (import (srfi 28))
> (format "Hello, ~a! You are ~a years old." "Alice" 30)
"Hello, Alice! You are 30 years old."
\end{lstlisting}

\subsection{SRFI-113: Sets and Bags}
\index{SRFI-113}\index{sets}

A set needs a \emph{comparator} (from SRFI-128) that tells it how to compare elements; \texttt{make-default-comparator} builds one that handles all the common types:

\begin{lstlisting}[style=repl]
> (import (srfi 113) (srfi 128) (srfi 132))
> (define s (set (make-default-comparator) 1 2 3 4 5))
> (set-contains? s 3)
#t
> (set-contains? s 9)
#f
> (list-sort < (set->list
                 (set-union s (set (make-default-comparator) 4 5 6 7))))
(1 2 3 4 5 6 7)
\end{lstlisting}

\texttt{set->list} returns the elements in no particular order, so the example sorts them for display.

\subsection{SRFI-189: Maybe and Either}
\index{SRFI-189}\index{maybe}\index{either}

\begin{lstlisting}[style=repl]
> (import (srfi 189))
> (define result (just 42))
> (maybe? result)
#t
> (maybe-ref result)
42
> (maybe-ref/default nothing 0)
0
\end{lstlisting}

\texttt{(just v)} wraps a present value; \texttt{nothing} is the singleton ``no value'' object.

\texttt{maybe} is like Python's \texttt{Optional} or Rust's \texttt{Option}---it represents a value that might not exist.

\section{Beyond SRFIs: Lazy Evaluation with \texttt{(scheme lazy)}}
\index{lazy evaluation}\index{delay}\index{force}\index{promises}

Not everything in the standard library is a SRFI. R7RS itself ships the \texttt{(scheme lazy)} library, which lets you defer a computation until its value is actually needed. \texttt{delay} wraps an expression into a \emph{promise} without evaluating it; \texttt{force} evaluates it on first use:

\begin{lstlisting}[style=repl]
> (import (scheme lazy))
> (define p (delay (begin (display "computing...")
                          (newline)
                          42)))
> (force p)
computing...
42
> (force p)
42
\end{lstlisting}

Note that \texttt{computing...} printed only once: promises \emph{memoize} their result, so the expression runs at most once no matter how often you force it. That makes \texttt{delay} a natural fit for expensive values that might never be needed---a parsed config file, a database connection, a large table.

Two companions round out the library. \texttt{make-promise} wraps an already-computed value so it can be used where a promise is expected, and \texttt{delay-force} is the tail-recursive variant of \texttt{delay} used when a promise's computation itself returns a promise:

\begin{lstlisting}[style=repl]
> (define q (make-promise 7))
> (force q)
7
> (promise? q)
#t
\end{lstlisting}

\begin{note}[Coming from Python]
A promise is like \texttt{functools.cache} applied to a zero-argument function: the first call computes, subsequent calls return the cached result. It is lazy initialization as a first-class value.
\end{note}

For programming with entire lazy sequences---infinite lists that materialize on demand---see SRFI-41 (streams) in Appendix~\ref{app:srfi}.

\section{Evaluating Data as Code: \texttt{(scheme eval)}}
\index{eval}\index{environments}\index{interaction-environment}

Chapter~\ref{ch:io} showed that \texttt{read} turns text into data structures without running anything. The \texttt{(scheme eval)} library supplies the other half: \texttt{eval} takes a data structure and evaluates it as code. Together they close the loop behind the phrase ``code is data''---programs are lists, and lists can become programs.

\texttt{eval} takes two arguments: the expression to evaluate and an \emph{environment} that supplies the bindings the expression may use. Build one with \texttt{environment}, which accepts the same import sets as an \texttt{import} form:

\begin{lstlisting}[style=repl]
> (import (scheme eval) (scheme repl))
> (eval '(+ 1 2) (environment '(scheme base)))
3
> (define form '(* 6 7))
> (eval form (environment '(scheme base)))
42
\end{lstlisting}

The expression is ordinary list data. You can assemble it with \texttt{list}, \texttt{cons}, or quasiquote (Chapter~\ref{ch:lists}) before handing it over---no string pasting involved.

An environment built by \texttt{environment} contains only the named libraries' bindings; your own definitions are not visible in it. To evaluate an expression against the live top-level environment, where everything you have defined is in scope, use \texttt{interaction-environment} from \texttt{(scheme repl)}:

\begin{lstlisting}[style=repl]
> (define x 10)
> (eval '(+ x 1) (interaction-environment))
11
\end{lstlisting}

Combining \texttt{read} on a string port (Chapter~\ref{ch:io}) with \texttt{eval} gives the closest analogue to Python's \texttt{eval("x + 5")}:

\begin{lstlisting}[style=repl]
> (define (eval-string s)
    (eval (read (open-input-string s))
          (interaction-environment)))
> (eval-string "(+ x 5)")
15
\end{lstlisting}

\begin{warning}[eval Runs Code]
\texttt{read} on untrusted input builds data at worst (Chapter~\ref{ch:io}); \texttt{eval} on untrusted input runs arbitrary code---file deletion included. Never feed \texttt{eval} anything that arrives over a network or from a user you do not control. And when the code is known as you write the program, you do not need \texttt{eval} at all: a procedure handles computed \emph{values}, and a macro (Chapter~\ref{ch:macros}) handles computed \emph{syntax}. Reach for \texttt{eval} only when program text genuinely arrives at run time---a configuration file of formulas, a calculator, a REPL of your own.
\end{warning}

\section{Clocks and Timing: \texttt{(scheme time)}}
\index{current-second}\index{current-jiffy}\index{jiffies-per-second}

R7RS also ships a minimal clock library, \texttt{(scheme time)}, with exactly three procedures. \texttt{current-second} returns the current time as a number of seconds since January 1, 1970---the same epoch as Python's \texttt{time.time()}. It is the right tool for timestamps, like the record timestamps in Chapter~\ref{ch:records}. For measuring \emph{durations}, use \texttt{current-jiffy}: a monotonically increasing tick count in implementation-defined units, with \texttt{jiffies-per-second} giving the conversion factor. In Kaappi a jiffy is a microsecond:

\begin{lstlisting}[style=repl]
> (import (scheme time))
> (jiffies-per-second)
1000000
\end{lstlisting}

A small helper turns any thunk into a timed run:

\begin{lstlisting}
(define (measure thunk)
  (let* ((start (current-jiffy))
         (result (thunk))
         (elapsed (- (current-jiffy) start)))
    (display "elapsed: ")
    (display (inexact (/ elapsed (jiffies-per-second))))
    (display "s")
    (newline)
    result))
\end{lstlisting}

\begin{lstlisting}
(measure (lambda () (length (iota 1000000))))
\end{lstlisting}

\begin{codeoutput}[Sample run]
elapsed: 0.021882s
\end{codeoutput}

The elapsed line differs on every run---this is a sample, not output to reproduce---and \texttt{measure} then returns the thunk's value, here \texttt{1000000}. Chapter~\ref{ch:macros} wrapped the same pattern in a \texttt{time-it} macro, so the timed expression does not need a \texttt{lambda}.

\section{Choosing the Right Data Structure}

\begin{center}
\begin{tabular}{llll}
\textbf{Need} & \textbf{Use} & \textbf{SRFI} & \textbf{Key operation} \\
\hline
Ordered sequence & List & 1 & O(n) search \\
Key-value map & Hash table & 69 & O(1) avg lookup \\
Sorted collection & Sorted list/vector & 132 & O(n log n) sort \\
Unique elements & Set & 113 & O(1) avg membership \\
FIFO queue & Queue & 117 & O(1) push/pop \\
Double-ended queue & Deque & 134 & O(1) at both ends \\
Indexed random access & Vector & 133 & O(1) index \\
\end{tabular}
\end{center}

\section*{Summary}

\begin{itemize}
    \item SRFIs are community-standard libraries; Kaappi ships 72 of them, each imported with \texttt{(import (srfi N))}.
    \item SRFI-1 extends list processing, SRFI-13 extends strings,
    and SRFI-69 provides hash tables.
    \item SRFI-133 (vectors), SRFI-132 (sorting), SRFI-27 (random
    numbers), and SRFI-115 (regular expressions) cover common needs.
    \item SRFI-158 provides generators, and other SRFIs add sets
    and bags, format strings, and the Maybe and Either types.
    \item SRFI-17 lets \texttt{set!} target accessor calls
    (\texttt{(set! (car p) v)}), and SRFI-61 adds a guard step
    to \texttt{cond}'s \texttt{=>} clauses.
    \item SRFI-64 is the standard test framework
    (\texttt{test-begin}, \texttt{test-end},
    \texttt{test-equal}, \texttt{test-assert},
    \texttt{test-error}); \texttt{kaappi test} runs whole suites.
    SRFI-78's \texttt{check} is a lightweight alternative.
    \item R7RS's own \texttt{(scheme lazy)} library provides
    \texttt{delay} and \texttt{force} for memoized, on-demand
    computation.
    \item \texttt{eval} evaluates list data as code, against
    either a library environment built by
    \texttt{environment} or the live
    \texttt{interaction-\allowbreak environment}. Never use it on
    untrusted input.
    \item \texttt{(scheme time)} provides
    \texttt{current-second} for wall-clock timestamps and
    \texttt{current-jiffy} with
    \texttt{jiffies-per-second} for measuring durations.
\end{itemize}

\section*{Exercises}

\begin{exercise}[Student Records with SRFI-1]
Given a list of student records represented as association lists:

\begin{lstlisting}
(define students
  '(("Alice" . 85) ("Bob" . 42) ("Carol" . 91)
    ("Dave" . 55)  ("Eve" . 73) ("Frank" . 38)))
\end{lstlisting}

Using SRFI-1 functions:
\begin{enumerate}
    \item Use \code{partition} to split the students into those who passed (score $\geq$ 60) and those who failed.
    \item Use \code{zip} to combine a list of names with a list of letter grades you assign.
    \item Use \code{unfold} to generate a list of threshold boundaries \code{(0 10 20 ... 100)} without manually writing them out.
\end{enumerate}
\end{exercise}

\begin{exercise}[Word Frequency Counter]
Write a procedure \code{(word-frequencies str)} that takes a string of words (separated by spaces) and returns a SRFI-69 hash table mapping each word to its count. Then write \code{(top-words ht n)} that returns the \code{n} most frequent words as a sorted list of \code{(word . count)} pairs.

\begin{lstlisting}[style=repl]
> (define ht (word-frequencies
    "the cat sat on the mat the cat"))
> (hash-table-ref ht "the")
3
> (hash-table-ref ht "cat")
2
> (top-words ht 2)
(("the" . 3) ("cat" . 2))
\end{lstlisting}

Hint: use \code{string-split} (SRFI-13) to tokenize
and \code{list-sort} (SRFI-132) for sorting. In Python
this would be \texttt{Counter(s.split())}.
\end{exercise}

\begin{exercise}[Email Validation with Regexps]
Using SRFI-115, write a procedure \code{(parse-email str)} that checks whether a string looks like an email address and extracts the local part and domain. Return \code{\#f} if the string is not a valid email, or a list \code{(local domain)} if it is.

\begin{lstlisting}[style=repl]
> (parse-email "alice@example.com")
("alice" "example.com")
> (parse-email "not-an-email")
#f
\end{lstlisting}

Use \code{regexp-matches} with submatch groups. A simple pattern is: one or more alphanumeric or dot/hyphen/underscore characters, then \code{@}, then one or more alphanumeric or dot/hyphen characters. You do not need to handle every edge case of RFC 5322.
\end{exercise}

\begin{exercise}[Fibonacci Generator]
Using SRFI-158, create a procedure \code{(make-fib-generator)} that returns a generator producing the Fibonacci sequence: 0, 1, 1, 2, 3, 5, 8, 13, \ldots\ Then write \code{(generator-take g n)}, which calls a generator \code{n} times and returns the collected values as a list.

\begin{lstlisting}[style=repl]
> (define fibs (make-fib-generator))
> (fibs)
0
> (fibs)
1
> (fibs)
1
> (fibs)
2
> (generator-take (make-fib-generator) 10)
(0 1 1 2 3 5 8 13 21 34)
\end{lstlisting}

Hint: use \code{make-unfold-generator} or build one manually with a closure that tracks two state variables. In Python, this would be a function with \code{yield}.
\end{exercise}

\begin{exercise}[Test Suite for Your Own Code]
Pick two exercises you solved earlier in this chapter (for example, \code{word-frequencies} and \code{parse-email}) and write a SRFI-64 test suite for them in a separate file. Cover the happy path, at least one edge case each (an empty string; a string with no \code{@}), and use \code{test-error} for an input that should raise an error. Run the file with \code{kaappi} and make every test pass. Then break one procedure on purpose, rerun the suite, and read the log file to see how the failure is reported.
\end{exercise}

In the next chapter, we explore calling C code from Scheme using the foreign function interface.

\chapterend
