\chapter{Concurrency}
\index{concurrency}\index{fibers}\index{threads}
\label{ch:concurrency}

Modern programs often need to do multiple things at once: serve multiple network connections, process data while waiting for I/O, or distribute work across CPU cores. Kaappi provides two concurrency models: \emph{fibers} (lightweight green threads) for cooperative multitasking and \emph{SRFI-18 threads} (OS threads) for true parallelism.

\section{Two Models of Concurrency}

\begin{center}
\begin{tabular}{lp{3.5cm}p{3.5cm}}
\textbf{Aspect} & \textbf{Fibers} & \textbf{OS Threads (SRFI-18)} \\
\hline
Library & \texttt{(kaappi fibers)} & \texttt{(srfi 18)} \\
Scheduling & Cooperative (yield) & Preemptive (OS) \\
Parallelism & No (single OS thread) & Yes (multiple cores) \\
Communication & Channels & Channels + deep-copy \\
Overhead & Low (kilobytes per fiber) & High (VM + GC per thread) \\
Shared state & Same heap & Separate heaps (safe) \\
Use case & I/O, event loops & CPU-bound parallelism \\
\end{tabular}
\end{center}

For most concurrent programs, fibers are the right choice. Use OS threads only when you need to exploit multiple CPU cores for compute-heavy work.

\begin{center}
\begin{tikzpicture}[
    fiber/.style={rectangle, draw, rounded corners, minimum width=3em, minimum height=2em,
                  font=\sffamily\small},
    chan/.style={rectangle, draw, thick, minimum width=2em, minimum height=1.5em,
                font=\sffamily\scriptsize},
    arr/.style={->, >=stealth, thick},
    lbl/.style={font=\sffamily\scriptsize}
]
\node[fiber] (f1) {Fiber 1};
\node[chan, right=1cm of f1] (ch) {channel};
\node[fiber, right=1cm of ch] (f2) {Fiber 2};
\node[fiber, below=0.6cm of f2] (f3) {Fiber 3};
\draw[arr] (f1) -- (ch) node[lbl, above, midway] {send};
\draw[arr] (ch) -- (f2) node[lbl, above, midway] {recv};
\draw[arr] (ch) |- (f3) node[lbl, right, near end] {recv};
\node[below=1.4cm of ch, font=\sffamily\scriptsize\itshape] {shared heap, cooperative scheduling};
\end{tikzpicture}
\end{center}

\section{Fibers}
\index{fibers}\index{spawn}

Fibers are lightweight green threads that run cooperatively within a single OS thread. They are created with \texttt{spawn}:

\begin{lstlisting}[style=repl]
> (import (kaappi fibers))
> (define f (spawn (lambda () (* 6 7))))
> (fiber-join f)
42
\end{lstlisting}

\texttt{spawn} takes a zero-argument procedure (a thunk), starts it as a new fiber, and returns a fiber object. \texttt{fiber-join}\index{fiber-join} blocks until the fiber completes and returns its result.

\subsection{Cooperative Scheduling}
\index{cooperative scheduling}

Fibers run until they voluntarily yield control. Call \texttt{yield}\index{yield} to let other fibers run:

\begin{lstlisting}[style=repl]
> (define (counting-fiber name n)
    (spawn (lambda ()
      (let loop ((i 0))
        (when (< i n)
          (display name) (display ":")
          (display i) (display " ")
          (yield)
          (loop (+ i 1)))))))
> (define f1 (counting-fiber "A" 3))
> (define f2 (counting-fiber "B" 3))
> (fiber-join f1)
A:0 B:0 A:1 B:1 A:2 B:2
> (fiber-join f2)
\end{lstlisting}

The fibers interleave at each \texttt{yield} point. Note that \texttt{spawn} only queues a fiber---nothing runs until the main fiber gives up control. The first \texttt{fiber-join} does exactly that, so both fibers run to completion there, and the second join returns immediately.

The flip side of cooperative scheduling is \emph{starvation}\index{starvation}. Here \texttt{slow} grinds through a long computation with no \texttt{yield} in sight, while \texttt{quick} only wants to print one line:

\begin{lstlisting}[style=repl]
> (define (busy-work n)
    (if (zero? n) 'done (busy-work (- n 1))))
> (define slow (spawn (lambda ()
    (busy-work 1000000)
    (display "slow finished") (newline))))
> (define quick (spawn (lambda ()
    (display "quick finished") (newline))))
> (fiber-join quick)
slow finished
quick finished
\end{lstlisting}

We asked for \texttt{quick}'s result, but \texttt{slow} was queued first and never gave up control, so \texttt{quick} had to wait out the entire computation. A \texttt{yield} in the loop restores responsiveness:

\begin{lstlisting}[style=repl]
> (define (busy-work n)
    (if (zero? n)
        'done
        (begin (yield) (busy-work (- n 1)))))
> (define slow (spawn (lambda ()
    (busy-work 1000000)
    (display "slow finished") (newline))))
> (define quick (spawn (lambda ()
    (display "quick finished") (newline))))
> (fiber-join quick)
quick finished
> (fiber-join slow)
slow finished
\end{lstlisting}

Now \texttt{slow} offers control back on every iteration, so \texttt{quick} runs at the first opportunity while the long computation finishes afterward. Be clear about what \texttt{yield} buys here: \emph{responsiveness}, not \emph{fairness}. It lets other fibers run; it does not promise an even split of processor time (the fan-out example in Section~\ref{sec:fan-out} returns to this).

\begin{warning}[Long computations without yield]
Since fibers are cooperative, a fiber that runs a long computation without yielding will block all other fibers. Insert \texttt{(yield)} in long loops to keep the system responsive.
\end{warning}

\subsection{Fiber Results and Errors}

\texttt{fiber-join} returns the fiber's result value. If the fiber raised an exception, \texttt{fiber-join} re-raises it in the calling context:

\begin{lstlisting}[style=repl]
> (define f (spawn (lambda () (error "oops"))))
> (guard (e (#t (error-object-message e)))
    (fiber-join f))
"oops"
\end{lstlisting}

\section{Channels}
\index{channels}\index{make-channel}\index{channel-send}\index{channel-receive}

Channels provide safe communication between fibers. A channel is a queue: \texttt{channel-send} appends a value and \texttt{channel-receive} takes the oldest one, letting other fibers run until a value is available. By default the queue is unbounded and a send returns immediately; \texttt{(make-channel n)} caps it at \texttt{n} queued values, and a send to a full bounded channel blocks the sending fiber until a receive frees a slot.

\begin{note}[Coming from Go]
Go's unbuffered channels synchronize sender and receiver: a send waits until a receive happens. Kaappi's default channel is closer to an unbounded buffered channel or an Erlang mailbox---producers never wait, so one that outruns its consumer grows the queue (and memory) with the backlog. A bounded channel \texttt{(make-channel n)} behaves like Go's \texttt{make(chan T, n)}: once \texttt{n} values are queued, senders wait---backpressure that keeps the producer honest. There is no zero-capacity rendezvous.
\end{note}

\begin{lstlisting}[style=repl]
> (import (kaappi fibers))
> (define ch (make-channel))
> (define sender
    (spawn (lambda ()
      (channel-send ch "hello from fiber"))))
> (channel-receive ch)
"hello from fiber"
\end{lstlisting}

\subsection{Bounded Channels: Backpressure}
\index{backpressure}\index{channels!bounded}

Give \texttt{make-channel} a capacity and the channel pushes back on a producer that gets ahead. Here the channel holds at most one value, so the producer's second send cannot complete until the consumer has taken the first:

\begin{lstlisting}[style=repl]
> (define ch (make-channel 1))
> (define producer
    (spawn (lambda ()
      (channel-send ch 'a)
      (display "sent a") (newline)
      (channel-send ch 'b)
      (display "sent b") (newline))))
> (channel-receive ch)
sent a
a
> (channel-receive ch)
sent b
b
> (fiber-join producer)
\end{lstlisting}

Follow the interleaving. The producer does not run until the first \texttt{channel-receive} blocks and hands over the scheduler; it sends \texttt{'a} into the free slot, prints, then blocks trying to send \texttt{'b} into the now-full channel. Each receive frees a slot and lets the producer take exactly one more step. That lock-step rhythm is the point: a bounded channel converts ``producer floods memory'' into ``producer waits its turn.''

\subsection{Timeouts and Closing}
\index{channel-close"!}\index{channels!timeouts}

Both \texttt{channel-receive} and \texttt{channel-send} accept an optional timeout in seconds, plus a value to return if it expires instead of raising an error:

\begin{lstlisting}[style=repl]
> (channel-receive (make-channel) 0.05 'nothing)
nothing
\end{lstlisting}

When a stream of values ends, close the channel. \texttt{channel-close!} marks it finished: receivers drain whatever is still queued and then get an EOF object---the same value a port returns at end of input (Chapter~\ref{ch:io})---while further sends raise an error. \texttt{channel-closed?} tests for it.

\begin{lstlisting}[style=repl]
> (define ch (make-channel))
> (channel-send ch 10)
> (channel-close! ch)
> (channel-receive ch)
10
> (channel-receive ch)
#<eof>
> (channel-send ch 99)
error[KP3000]: channel-send: send on closed channel
\end{lstlisting}

One more safety net: if a receive (or a \texttt{fiber-join}) can never be satisfied because every other fiber is blocked or finished, Kaappi does not hang. It raises a catchable error naming the deadlock. You will meet it the first time you forget to close a channel or send a sentinel---read it as a precise bug report, not a nuisance.

\subsection{Producer-Consumer Pattern}
\index{producer-consumer}

\begin{lstlisting}
(import (kaappi fibers))

(define ch (make-channel))

;; Producer: sends numbers 0-9
(define producer
  (spawn (lambda ()
    (let loop ((i 0))
      (when (< i 10)
        (channel-send ch i)
        (loop (+ i 1))))
    (channel-send ch 'done))))

;; Consumer: receives until 'done
(let loop ()
  (let ((msg (channel-receive ch)))
    (unless (eq? msg 'done)
      (display (* msg msg))
      (display " ")
      (loop))))
\end{lstlisting}

\begin{codeoutput}
0 1 4 9 16 25 36 49 64 81
\end{codeoutput}

\subsection{Fan-Out: Multiple Workers}
\label{sec:fan-out}

\begin{lstlisting}
(import (kaappi fibers))

(define work-ch (make-channel))
(define result-ch (make-channel))

;; Spawn 3 worker fibers
(define workers
  (map (lambda (id)
    (spawn (lambda ()
      (let loop ()
        (let ((task (channel-receive work-ch)))
          (unless (eq? task 'stop)
            (channel-send result-ch
              (list id (* task task)))
            (loop)))))))
    '(1 2 3)))

;; Send work, then one stop sentinel per worker
(for-each (lambda (n) (channel-send work-ch n))
          '(10 20 30 40 50 60))
(for-each (lambda (w) (channel-send work-ch 'stop))
          workers)

;; Collect results, then wait for the workers to exit
(define results
  (let loop ((n 6) (acc '()))
    (if (= n 0)
        acc
        (loop (- n 1)
              (cons (channel-receive result-ch) acc)))))
(for-each fiber-join workers)
(display results)
\end{lstlisting}

\begin{codeoutput}
((1 3600) (1 2500) (1 1600) (1 900) (1 400) (1 100))
\end{codeoutput}

Look at the worker ids: every result came from worker 1. With cooperative scheduling and non-blocking sends, the first worker to run keeps going until it blocks---and since the whole queue is already buffered, it drains everything before workers 2 and 3 get a turn. Fan-out pays off when each task genuinely blocks (waiting on a socket, a timer, or another channel), which hands the queue to the next worker. Always send one stop sentinel per worker: a worker that never receives one stays blocked on the channel forever.

You might hope that an explicit \texttt{(yield)} inside the worker loop would spread the work evenly. Try it: depending on where you place the call, worker 1 still takes every task, or at best hands one task each to the other workers before draining the rest. The scheduler promises only that \emph{some} runnable fiber runs next---it makes no fairness guarantee, and \texttt{yield} is a hint, not a turnstile. If your tasks are pure computation and you need them spread across workers, fibers are the wrong tool; use OS threads (next section).

\subsection{Racing Fibers}

You can race multiple fibers by having them send to a shared channel. The first result wins:

\begin{lstlisting}
(define (race thunk1 thunk2)
  (let ((ch (make-channel)))
    (spawn (lambda () (channel-send ch (thunk1))))
    (spawn (lambda () (channel-send ch (thunk2))))
    (channel-receive ch)))
\end{lstlisting}

\begin{lstlisting}[style=repl]
> (race (lambda () (+ 1 2))
        (lambda () (* 3 4)))
3
\end{lstlisting}

The result is whichever fiber completes first. The other fiber's value is left in the channel (a second \texttt{channel-receive} would retrieve it). Note that with cooperative scheduling, two thunks that never block run to completion in spawn order---the first one always wins. Racing becomes meaningful when the thunks block on I/O, timers, or other channels.

\section{OS Threads (SRFI-18)}
\index{SRFI-18}\index{OS threads}

For true parallelism across CPU cores, use SRFI-18 OS threads. Each thread gets its own VM and garbage collector with an independent heap.

\begin{warning}[Heavyweight concurrency]
Each OS thread creates an independent VM and garbage collector. This makes them safe (no shared mutable state) but expensive. Prefer fibers for I/O concurrency; reserve threads for CPU-bound work that benefits from multiple cores. OS threads are not available in sandbox mode or WASM builds.
\end{warning}

\begin{lstlisting}
(import (srfi 18))

(define t (thread-start!
  (make-thread
    (lambda () (* 6 7)))))

(thread-join! t)  ;=> 42
\end{lstlisting}

\subsection{Creating and Starting Threads}
\index{make-thread}\index{thread-start"!}\index{thread-join"!}

\begin{lstlisting}
(import (srfi 18))

(define (expensive-computation n)
  (let loop ((i 0) (acc 0))
    (if (= i n) acc
        (loop (+ i 1) (+ acc i)))))

(define t1 (thread-start!
  (make-thread
    (lambda () (expensive-computation 10000000))
    "worker-1")))

(define t2 (thread-start!
  (make-thread
    (lambda () (expensive-computation 20000000))
    "worker-2")))

(define result1 (thread-join! t1))
(define result2 (thread-join! t2))
\end{lstlisting}

Both computations run in parallel on separate cores.

\subsection{Deep-Copy Semantics}

Values crossing thread boundaries are \emph{deep-copied}:

\begin{itemize}
    \item At \texttt{thread-start!}: the thunk closure (and all values it captures) is deep-copied from the parent thread to the child.
    \item At \texttt{thread-join!}: the return value is deep-copied from the child thread back to the parent.
\end{itemize}

This means threads cannot share mutable state---there are no race conditions, no data corruption, and no need for locks in most code. The one sanctioned shared object is a channel, covered next.

\begin{note}[Coming from Python]
Python's \texttt{multiprocessing} module works similarly: separate memory spaces, data copied between processes. Kaappi's threads are lighter (same process, shared code) but have the same safety guarantee of no shared mutable state.
\end{note}

\subsection{Channels Across Threads}
\index{channels!across threads}

Waiting for \texttt{thread-join!} is not the only way to hear back from a thread. A channel handed to the thread's thunk is \emph{promoted} to a shared, thread-safe channel, and both sides talk through it while the thread runs---results can stream back long before the join:

\begin{lstlisting}
(import (kaappi fibers) (srfi 18))

(define (start-worker out)
  (thread-start!
    (make-thread
      (lambda ()
        (channel-send out (* 6 7))))))

(define ch (make-channel))
(define t (start-worker ch))
(channel-receive ch)  ;=> 42
(thread-join! t)
\end{lstlisting}

Each value sent across still crosses by copy, so the no-shared-mutable-state guarantee survives; the channel itself is the only object both threads touch. Everything from the previous sections works across threads too: capacities give you backpressure between threads, and \texttt{channel-close!} signals end-of-stream.

\begin{warning}[Hand the channel over explicitly]
A thread may only use a channel that was \emph{handed} to it---captured by the thunk as a parameter or a \texttt{let}-bound local, as \texttt{out} is above. Reaching a channel through a global variable from inside the thunk does not promote it: the child's send fails and the parent's receive deadlocks. When a thread needs a channel, pass it in.
\end{warning}

\subsection{Mutexes and Condition Variables}
\index{mutex}\index{condition variables}

SRFI-18 provides mutexes and condition variables for the rare cases where you need synchronization:

\begin{lstlisting}
(import (srfi 18))

(define mtx (make-mutex))

(mutex-lock! mtx)
;; ... critical section ...
(mutex-unlock! mtx)
\end{lstlisting}

In practice, the deep-copy model means you rarely need mutexes. Use them only if you are integrating with external resources (files, network sockets) that require coordinated access.

\section{Choosing Between Fibers and Threads}

\begin{center}
\begin{tabular}{llp{3.2cm}}
\textbf{Scenario} & \textbf{Use} & \textbf{Why} \\
\hline
Multiple HTTP connections & Fibers & I/O-bound, many tasks \\
Many connections across cores & Threads $\times$ fibers & \texttt{http-listen-}\allowbreak\texttt{parallel} \\
Event loop / GUI & Fibers & Cooperative scheduling \\
Pipeline processing & Fibers + channels & Natural producer-consumer \\
CPU-heavy computation & OS threads & True parallelism \\
Parallel map over data & Pooled threads & \texttt{parallel-map} \\
Simple background task & Fibers & Lower overhead \\
\end{tabular}
\end{center}

\begin{note}[Why fibers suit I/O-bound work]
A blocking-looking call like \texttt{(read-line port)} only suspends the fiber that made it, not the whole program. Kaappi runs a per-OS-thread I/O reactor: when a read or write would block, the calling fiber parks and the reactor waits on the underlying file descriptor, waking the fiber once data is ready. \texttt{thread-sleep!} gets the same treatment, parking on a timer instead of blocking the thread. This is why a fiber-based server can hold thousands of connections open on a single OS thread and GC heap---a slow connection parks its own fiber without freezing the rest. The ecosystem's \texttt{http-listen-fiber} (Appendix~\ref{app:ecosystem}) is a fiber-per-connection HTTP server built on exactly this. To use every core rather than one, \texttt{http-listen-parallel} runs one such fiber server per OS thread; on Linux the kernel spreads incoming connections across the threads' \texttt{SO\_REUSEPORT} sockets, so the machine serves \emph{threads $\times$ fibers}---cores times thousands of connections. (macOS does not balance \texttt{SO\_REUSEPORT}, so parallel serving there falls back to one acceptor thread feeding a worker pool over a shared channel; each worker still fiber-multiplexes its connections, so macOS reaches the same \emph{threads $\times$ fibers} model with a userspace distributor. Linux stays the faster path, but both scale across cores.)
\end{note}

\section{A Complete Example: Parallel Map}

Here is a function that maps a procedure over a list using OS threads for parallelism:

\begin{lstlisting}
(import (srfi 18))

(define (parallel-map f lst)
  (let ((threads
          (map (lambda (item)
                 (thread-start!
                   (make-thread
                     (lambda () (f item)))))
               lst)))
    (map thread-join! threads)))
\end{lstlisting}

\begin{lstlisting}[style=repl]
> (parallel-map (lambda (n) (* n n)) '(1 2 3 4 5))
(1 4 9 16 25)
\end{lstlisting}

Each element is processed on a separate thread. For CPU-intensive \texttt{f}, this achieves near-linear speedup on multi-core machines.

\section{The \texttt{(kaappi parallel)} Library}
\index{kaappi parallel}\index{parallel-map}\index{make-pool}\index{processor-count}

The pattern above---one thread per element, join in order---is common enough that Kaappi ships a library for it. \texttt{(kaappi parallel)} provides \texttt{parallel-map} and \texttt{parallel-for-each}, built on a fixed pool of worker threads draining a shared task channel:

\begin{lstlisting}[style=repl]
> (import (kaappi parallel))
> (parallel-map (lambda (n) (* n n)) '(1 2 3 4 5))
(1 4 9 16 25)
\end{lstlisting}

For finer control, build the pool yourself. \texttt{(make-pool n)} starts \texttt{n} worker threads; \texttt{pool-submit} hands one a thunk and immediately returns a task handle; \texttt{task-wait} collects that task's result; \texttt{pool-shutdown!} stops the workers:

\begin{lstlisting}[style=repl]
> (define pool (make-pool 2))
> (define task
    (pool-submit pool (lambda () (+ 40 2))))
> (task-wait task)
42
> (pool-shutdown! pool)
\end{lstlisting}

\texttt{(processor-count)} reports how many CPU cores the machine has---the natural pool size. Under \texttt{--sandbox} and on WASM, where OS threads do not exist, the same API degrades gracefully to fiber workers: your program still runs, just without the parallelism.

\begin{note}[One task per element has a cost]
\texttt{parallel-map} submits one pool task per list element, so every element pays a cross-thread round trip; it suits lists up to a few hundred elements whose per-element work is substantial. For bigger inputs, chunk: submit one \texttt{pool-submit} task per core, each covering a slice of the input with an ordinary sequential loop. The tasks-as-envelopes model is the same one \texttt{thread-start!} uses, so thunks and results cross by copy here too.
\end{note}

\section{A Complete Example: Fiber Pipeline}

A pipeline where data flows through multiple stages via channels. The producer and the transformer each run as a fiber; the main fiber acts as the final stage. An \texttt{'eof} sentinel flows through the channels to shut the pipeline down:

\begin{lstlisting}
(import (kaappi fibers))

(define nums (make-channel))
(define squares (make-channel))

;; Stage 1: produce 1..5
(define producer
  (spawn (lambda ()
    (for-each (lambda (n) (channel-send nums n))
              '(1 2 3 4 5))
    (channel-send nums 'eof))))

;; Stage 2: square each value
(define squarer
  (spawn (lambda ()
    (let process ()
      (let ((val (channel-receive nums)))
        (if (eq? val 'eof)
            (channel-send squares 'eof)
            (begin
              (channel-send squares (* val val))
              (process))))))))
\end{lstlisting}

\begin{lstlisting}
;; Stage 3: add 1 and print (the main fiber)
(let loop ()
  (let ((val (channel-receive squares)))
    (unless (eq? val 'eof)
      (display (+ val 1)) (display " ")
      (loop))))
\end{lstlisting}

\begin{codeoutput}
2 5 10 17 26
\end{codeoutput}

Each number is squared by one fiber and incremented by the next stage, with the channels handing values downstream one at a time.

The \texttt{'eof} sentinel is the hand-rolled version of what \texttt{channel-close!} provides: close a channel after the last send, and once the queue drains every receive returns an EOF object. The same producer, with the consumer testing \texttt{eof-object?} instead of comparing against a sentinel:

\begin{lstlisting}
(define nums2 (make-channel))

(define producer2
  (spawn (lambda ()
    (for-each (lambda (n) (channel-send nums2 n))
              '(1 2 3 4 5))
    (channel-close! nums2))))

(let loop ()
  (let ((val (channel-receive nums2)))
    (unless (eof-object? val)
      (display (* val val)) (display " ")
      (loop))))
\end{lstlisting}

\begin{codeoutput}
1 4 9 16 25
\end{codeoutput}

Closing scales better than a sentinel when a channel has several readers: every blocked receiver wakes and sees EOF, whereas a sentinel is consumed by whichever single reader happens to take it (the fan-out example sent one \texttt{'stop} per worker for exactly that reason).

\section*{Summary}

\begin{itemize}
    \item Kaappi offers two concurrency models: lightweight cooperative fibers and preemptive OS threads.
    \item Fibers are scheduled cooperatively and communicate through channels---unbounded by default, bounded for backpressure---supporting producer-consumer, fan-out, and racing patterns.
    \item Channel sends and receives take optional timeouts; \code{channel-close!} signals end-of-stream (receivers drain, then get an EOF object); an unsatisfiable wait raises a deadlock error instead of hanging.
    \item OS threads (SRFI-18) use deep-copy semantics and synchronize with mutexes and condition variables; a channel handed to a thread's thunk is promoted and shared, so threads can stream results back before the join.
    \item \code{(kaappi parallel)} adds \code{parallel-map} and worker pools (\code{make-pool}, \code{pool-submit}, \code{task-wait}) over pooled OS threads.
    \item Use fibers for I/O-bound concurrency and threads for CPU-bound parallelism.
\end{itemize}

\section*{Exercises}

\begin{exercise}[Three-Stage Fiber Pipeline]
Write a three-stage fiber pipeline connected by channels: a \emph{producer} that sends the integers 1 through 10, a \emph{transformer} that squares each value, and a \emph{consumer} that collects the results into a list. Use a sentinel value \code{'done} to signal the end of the stream.

\begin{lstlisting}
(import (kaappi fibers))

(define ch1 (make-channel))  ; producer -> transformer
(define ch2 (make-channel))  ; transformer -> consumer

;; Producer: sends 1..10 then 'done
(spawn (lambda () ...))

;; Transformer: reads from ch1, squares, sends to ch2
(spawn (lambda () ...))

;; Consumer: reads from ch2 until 'done
;; Should produce: (1 4 9 16 25 36 49 64 81 100)
\end{lstlisting}

This is the Scheme equivalent of a Unix pipeline like \texttt{seq 10 | awk '\{print \$1*\$1\}'}.
\end{exercise}

\begin{exercise}[Fiber-Based Parallel Map]
Implement \code{(fiber-parallel-map f lst)} using fibers and channels. Spawn one fiber per element, have each fiber send its result to a shared results channel, then collect all results in order.

\begin{lstlisting}[style=repl]
> (fiber-parallel-map
    (lambda (x) (* x x))
    '(1 2 3 4 5))
(1 4 9 16 25)
\end{lstlisting}

Hint: to preserve order, have each fiber send a pair \code{(index . result)} and sort by index after collecting. Alternatively, use one channel per fiber. How does this differ from the OS-thread \code{parallel-map} shown earlier in the chapter? (Think about whether fibers actually compute in parallel.)
\end{exercise}

\begin{exercise}[Collecting Results from N Fibers]
Write a procedure \code{(spawn-collect thunks)} that takes a list of zero-argument procedures, spawns each as a fiber, and returns a list of all their results. Use a single shared channel for collection.

\begin{lstlisting}[style=repl]
> (spawn-collect
    (list (lambda () (+ 1 2))
          (lambda () (* 3 4))
          (lambda () (string-length "hello"))))
(3 12 5)
\end{lstlisting}

The results may arrive in any order---your procedure should still return them in the same order as the input thunks. This pattern is common when you need to fan-out work and wait for all results, similar to Python's \code{asyncio.gather()}.
\end{exercise}

\begin{exercise}[Worker Pool (Challenge)]
Implement a worker pool with a fixed number of worker fibers that process jobs from a shared channel:

\begin{lstlisting}
(define (make-worker-pool num-workers job-proc)
  ;; Returns two values: a job channel and a
  ;; results channel. Spawns num-workers fibers
  ;; that each loop reading from the job channel,
  ;; applying job-proc, and sending the result to
  ;; the results channel.
  ...)
\end{lstlisting}

Have each worker exit its loop when it receives \code{'stop}, and send one \code{'stop} per worker once all jobs are submitted. Test the pool with 3 workers and 10 jobs (the harness needs \code{(srfi 132)} for \code{list-sort}):

\begin{lstlisting}
(let-values (((jobs results)
              (make-worker-pool 3
                (lambda (n) (* n n)))))
  ;; Submit 10 jobs, then stop the workers
  (for-each (lambda (n) (channel-send jobs n))
            '(1 2 3 4 5 6 7 8 9 10))
  (for-each (lambda (i) (channel-send jobs 'stop))
            '(1 2 3))
  ;; Collect 10 results
  (let loop ((n 10) (acc '()))
    (if (= n 0) (list-sort < acc)
        (loop (- n 1)
              (cons (channel-receive results)
                    acc)))))
\end{lstlisting}

Why would you use a fixed pool of workers instead of spawning one fiber per job? (Hint: think about what happens with 10,000 jobs that each open a network connection.) This is analogous to Python's \code{concurrent.futures.ThreadPoolExecutor}---and to \code{make-pool} from \code{(kaappi parallel)}, which applies the same design across OS threads.
\end{exercise}

This completes Part~IV. The appendices that follow provide quick references for the R7RS standard and the Kaappi CLI.

\chapterend
